---
title: "Rcode_manuscript"
author: "Mengyi Dong"
date: "2026-05-07"
output: html_document
---

#Load packages
```{r}
library(phyloseq)
library(FUNGuildR)
library(Biostrings)
library(qiime2R)
library(tidyr)
library(dplyr)
library(ape)
library(vegan)
library(tibble)
library(ggplot2)
library(stringr)
library(tibble)
library(ggpubr)
library(ggExtra)
library(ggalluvial)
```

#Load data
```{r}
#The PacBio fungi dataset with only the samples matched Illumina ITS2 dataset
ps_PacBio_Fungi_match <- readRDS("PacBio/ps_PacBio_Fungi_match_sklearn.rds")
tax_ITS <- data.frame(ps_PacBio_Fungi_match@tax_table)
metadata_ITS <- data.frame(ps_PacBio_Fungi_match@sam_data)
otu_ITS <- data.frame(ps_PacBio_Fungi_match@otu_table)

#The PacBio fungi dataset trimmed to ITS2 region with only samples match Illumina ITS2 dataset
ps_PacBio_ITS2_match <- readRDS("PacBio/ps_PacBio_ITS2_match.rds")
tax_pacbio_ITS2 <- data.frame(ps_PacBio_ITS2_match@tax_table)
metadata_pacbio_ITS2 <- data.frame(ps_PacBio_ITS2_match@sam_data)
otu_pacbio_ITS2 <- data.frame(ps_PacBio_ITS2_match@otu_table)

#The cleaned Illumina ITS2 dataset
ps_Illumina_ITS2 <- readRDS("Illumina/ps_Illumina_ITS2_sklearn.rds")
tax_ITS2 <- data.frame(ps_Illumina_ITS2@tax_table)
metadata_ITS2 <- data.frame(ps_Illumina_ITS2@sam_data)
otu_ITS2 <- data.frame(ps_Illumina_ITS2@otu_table)
```

#Rarefaction
##Fig S2: determine where a plateau occurs for each dataset
```{r}
# Convert to matrix, make sure OTUs are in rows for vegan
otu_ITS <- t(as(otu_table(ps_PacBio_Fungi_match), "matrix"))
otu_ITS2 <- t(as(otu_table(ps_Illumina_ITS2), "matrix"))
otu_ITS2_pacbio <- t(as(otu_table(ps_PacBio_ITS2_match), "matrix"))

otu_ITS2_filt <- otu_ITS2[rowSums(otu_ITS2, na.rm = TRUE) > 1, , drop = FALSE]

# Plot rarefaction curves
rarecurve(otu_ITS, step = 1000, sample=min(rowSums(otu_ITS)), label = FALSE, col = "deeppink", main = "PacBio ITS Rarefaction")
rarecurve(otu_ITS2_filt, step = 1000, sample=min(rowSums(otu_ITS2)),label = FALSE, col = "orange", main = "Illumina ITS2 Rarefaction")
rarecurve(otu_ITS2_pacbio, step = 1000, sample=min(rowSums(otu_ITS2_pacbio)),label = FALSE, col = "deeppink", main = "Trimmed ITS2 Rarefaction")
```

##Choose a Rarefaction Depth
```{r}
summary(sample_sums(ps_PacBio_Fungi_match))
summary(sample_sums(ps_Illumina_ITS2))

depth_shared <- 2000  # Choose based on your previous rarefaction curve

# Rarefy both to same depth
set.seed(123)
ps_PacBio_rarefied <- rarefy_even_depth(ps_PacBio_Fungi_match, sample.size = depth_shared, rngseed = 123, verbose = FALSE)
ps_Illumina_rarefied <- rarefy_even_depth(ps_Illumina_ITS2, sample.size = depth_shared, rngseed = 123, verbose = FALSE)

ps_PacBio_rarefied
ps_Illumina_rarefied

# Number of non-zero taxa per sample = ASVs detected
asv_counts_PacBio <- apply(otu_table(ps_PacBio_rarefied), 2, function(x) sum(x > 0))
asv_counts_Illumina <- apply(otu_table(ps_Illumina_rarefied), 2, function(x) sum(x > 0))

# Convert both vectors to data frames
common_samples <- intersect(names(asv_counts_PacBio), names(asv_counts_Illumina))


asv_df <- data.frame(
  SampleID = common_samples,
  ASVs_PacBio = asv_counts_PacBio[common_samples],
  ASVs_Illumina = asv_counts_Illumina[common_samples]
)

#Plot for comparison
summary_ASV_detected <- ggplot(asv_df, aes(x = ASVs_Illumina, y = ASVs_PacBio)) +
  geom_point(size = 3, alpha = 0.3, color = "black") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )+
  labs(x = "ASVs (Short-read)", y = "ASVs (Long-read)", title = paste("ASV Richness Comparison at", depth_shared, "Reads"))

summary_ASV_detected

#Visualize paired difference
asv_long <- asv_df %>%
  pivot_longer(cols = starts_with("ASVs"), names_to = "Pipeline", values_to = "ASV_Count")

paired <- ggplot(asv_long, aes(x = Pipeline, y = ASV_Count, group = SampleID)) +
  geom_line(alpha = 0.3) +
  geom_point(size = 3, alpha = 0.3) +
  labs(title = "ASVs Detected per Sample by Pipeline", y = "Number of ASVs", x = "Pipeline") +
  theme_bw(base_size = 14)+
   theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )

paired
```


#Good's coverage ITS2
```{r}
#Good’s coverage = 1 - (singletons / total reads) per sample
goods_coverage <- function(ps) {
  mat <- as(otu_table(ps), "matrix")
  if (!taxa_are_rows(ps)) mat <- t(mat)
  # counts: taxa x samples
  singles <- colSums(mat == 1)
  totals  <- colSums(mat)
  tibble(sample_id = colnames(mat),
         goods = 1 - (singles / totals))
}

gc_its2 <- goods_coverage(ps_Illumina_ITS2)
summary(gc_its2$goods)

gc_its <- goods_coverage(ps_PacBio_Fungi_match)
summary(gc_its$goods)

#Low coverage outlier
low_cov <- gc_its2 %>% arrange(goods) %>% head(3)
low_cov

# decide a threshold (e.g., < 0.90); report n low-coverage samples
sum(gc_its2$goods < 0.90, na.rm = TRUE)
sum(gc_its$goods < 0.90, na.rm = TRUE)

# Optional: drop extremely low-coverage samples from ITS2 *and* their paired PacBio samples
drop_ids <- gc_its2 %>% filter(goods < 0.90) %>% pull(sample_id)
drop_ids
ps_Illumina_ITS2_filt <- prune_samples(!(sample_names(ps_Illumina_ITS2) %in% drop_ids), ps_Illumina_ITS2)
ps_Illumina_ITS2_filt <- prune_taxa(taxa_sums(ps_Illumina_ITS2_filt) > 0, ps_Illumina_ITS2_filt)
ps_PacBio_ITS_filt <- prune_samples(sample_names(ps_Illumina_ITS2_filt), ps_PacBio_Fungi_match)
ps_PacBio_ITS_filt <- prune_taxa(taxa_sums(ps_PacBio_ITS_filt) > 0, ps_PacBio_ITS_filt)
ps_PacBio_ITS2_filt <- prune_samples(sample_names(ps_Illumina_ITS2_filt), ps_PacBio_ITS2_match)
ps_PacBio_ITS2_filt <- prune_taxa(taxa_sums(ps_PacBio_ITS2_filt) > 0, ps_PacBio_ITS2_filt)

#Report coverage
n_total <- nrow(gc_its2)
n_ge95  <- sum(gc_its2$goods >= 0.95, na.rm = TRUE)
paste0(n_ge95, "/", n_total, " (", round(100*n_ge95/n_total), "%) of ITS2 samples ≥ 0.95 coverage.")
```

#Create subsets
## Subsetting by SampleType: Illumina ITS2
```{r}
ps <- ps_Illumina_ITS2_filt
ps@sam_data$Reads <- sample_sums(ps)
metadata_ITS2 <- data.frame(ps@sam_data)

ps_ITS2_Control <- subset_samples(ps, SampleType %in% c("NegCon","PosCon"))
ps_ITS2_Control <- prune_taxa(taxa_sums(ps_ITS2_Control) > 0, ps_ITS2_Control)
ps_ITS2_Control

ps_ITS2_PosCon <- subset_samples(ps, SampleType == "PosCon")
ps_ITS2_PosCon <- prune_taxa(taxa_sums(ps_ITS2_PosCon) > 0, ps_ITS2_PosCon)
ps_ITS2_PosCon

ps_ITS2_ntc <- subset_samples(ps, Type == "ntc")
ps_ITS2_ntc <- prune_taxa(taxa_sums(ps_ITS2_ntc)>0, ps_ITS2_ntc)
ps_ITS2_ntc

ps_ITS2_nc <- subset_samples(ps, Type == "nc")
ps_ITS2_nc <- prune_taxa(taxa_sums(ps_ITS2_nc)>0, ps_ITS2_nc)
ps_ITS2_nc

ps_ITS2_nec <- subset_samples(ps, Type == "nec")
ps_ITS2_nec <- prune_taxa(taxa_sums(ps_ITS2_nec)>0, ps_ITS2_nec)
ps_ITS2_nec

ps_ITS2_MB <- subset_samples(ps, Type == "method_blank")
ps_ITS2_MB <- prune_taxa(taxa_sums(ps_ITS2_MB)>0, ps_ITS2_MB)
ps_ITS2_MB

ps_ITS2_FB <- subset_samples(ps, Type == "field_blank")
ps_ITS2_FB <- prune_taxa(taxa_sums(ps_ITS2_FB)>0, ps_ITS2_FB)
ps_ITS2_FB

ps_ITS2_Sample <- subset_samples(ps, SampleType %in% c("air","swab"))
ps_ITS2_Sample <- prune_taxa(taxa_sums(ps_ITS2_Sample) > 0, ps_ITS2_Sample)
ps_ITS2_Sample

ps_ITS2_AirSwab <- subset_samples(ps, Location %in% c("Home1","Lab1","Lab2"))
ps_ITS2_AirSwab <- prune_taxa(taxa_sums(ps_ITS2_AirSwab) > 0, ps_ITS2_AirSwab)
ps_ITS2_AirSwab
```

## Subsetting by SampleType: PacBio ITS
```{r}
ps <- ps_PacBio_ITS_filt
ps@sam_data$Reads <- sample_sums(ps)
metadata_ITS <- data.frame(ps@sam_data)

ITS_control <- subset_samples(ps, SampleType %in% c("NegCon","PosCon"))
ITS_control <- prune_taxa(taxa_sums(ITS_control)>0, ITS_control)
ITS_control

ITS_PosCon <- subset_samples(ps, SampleType == "PosCon" )
ITS_PosCon <- prune_taxa(taxa_sums(ITS_PosCon)>0, ITS_PosCon)
ITS_PosCon

ITS_ntc <- subset_samples(ps, Type == "ntc")
ITS_ntc <- prune_taxa(taxa_sums(ITS_ntc)>0, ITS_ntc)
ITS_ntc

ITS_nec <- subset_samples(ps, Type == "nec")
ITS_nec <- prune_taxa(taxa_sums(ITS_nec)>0, ITS_nec)
ITS_nec

ITS_MB <- subset_samples(ps, Type == "method_blank")
ITS_MB <- prune_taxa(taxa_sums(ITS_MB)>0, ITS_MB)
ITS_MB

ITS_FB <- subset_samples(ps, Type == "field_blank")
ITS_FB <- prune_taxa(taxa_sums(ITS_FB)>0, ITS_FB)
ITS_FB

ps_ITS_sample <- subset_samples(ps, SampleType %in% c("air","swab"))
ps_ITS_sample <- prune_taxa(taxa_sums(ps_ITS_sample)>0, ps_ITS_sample)
ps_ITS_sample

ps_ITS_AirSwab <- subset_samples(ps, Location %in% c("Home1","Lab1","Lab2"))
ps_ITS_AirSwab <- prune_taxa(taxa_sums(ps_ITS_AirSwab) > 0, ps_ITS_AirSwab)
ps_ITS_AirSwab
```

## Merge same sequences PacBio ITS2
```{r}
seqs <- as.character(refseq(ps_PacBio_ITS2_filt))
table(duplicated(seqs))

dup_seqs <- seqs[duplicated(seqs) | duplicated(seqs, fromLast = TRUE)]
length(unique(dup_seqs))

dup_df <- data.frame(
  ASV = taxa_names(ps_PacBio_ITS2_filt),
  Sequence = seqs,
  stringsAsFactors = FALSE
)

# Extract otu table
otu_mat <- as(otu_table(ps_PacBio_ITS2_filt), "matrix")

# Make sure taxa are rows
if (!taxa_are_rows(ps_PacBio_ITS2_filt)) {
  otu_mat <- t(otu_mat)
}

# Extract sequences
seqs <- as.character(refseq(ps_PacBio_ITS2_filt))

# Sum counts for identical sequences
otu_merged <- rowsum(otu_mat, group = seqs)

# Build merged tax table
tax_mat <- as(tax_table(ps_PacBio_ITS2_filt), "matrix")
tax_df <- data.frame(
  Sequence = seqs,
  tax_mat,
  stringsAsFactors = FALSE
)

# Keep first taxonomy entry per unique sequence
# If taxonomy differs among collapsed ASVs, see note below
tax_df_merged <- tax_df[!duplicated(tax_df$Sequence), ]
rownames(tax_df_merged) <- tax_df_merged$Sequence
tax_df_merged$Sequence <- NULL

# Keep rows in same order as merged otu table
tax_df_merged <- tax_df_merged[rownames(otu_merged), , drop = FALSE]

# Build merged refseq object
refseq_merged <- Biostrings::DNAStringSet(rownames(otu_merged))
names(refseq_merged) <- rownames(otu_merged)

# Rebuild phyloseq
ps_trimmed_ITS2_merged <- phyloseq(
  otu_table(otu_merged, taxa_are_rows = TRUE),
  tax_table(as.matrix(tax_df_merged)),
  sample_data(ps_PacBio_ITS2_filt),
  refseq(refseq_merged)
)

ntaxa(ps_PacBio_ITS2_filt)
ntaxa(ps_trimmed_ITS2_merged)

seqs2 <- as.character(refseq(ps_trimmed_ITS2_merged))
any(duplicated(seqs2))
```

## Subsetting by SampleType: Trimmed ITS2
```{r}
ps <- ps_PacBio_ITS2_filt

pacbio_ITS2_control <- subset_samples(ps, SampleType %in% c("NegCon","PosCon"))
pacbio_ITS2_control <- prune_taxa(taxa_sums(pacbio_ITS2_control)>0, pacbio_ITS2_control)
pacbio_ITS2_control

pacbio_ITS2_sample <- subset_samples(ps, SampleType  %in% c("air","swab"))
pacbio_ITS2_sample <- prune_taxa(taxa_sums(pacbio_ITS2_sample)>0, pacbio_ITS2_sample)
pacbio_ITS2_sample

pacbio_ITS2_AirSwab <- subset_samples(ps, Location %in% c("Home1","Lab1","Lab2"))
pacbio_ITS2_AirSwab <- prune_taxa(taxa_sums(pacbio_ITS2_AirSwab) > 0, pacbio_ITS2_AirSwab)
pacbio_ITS2_AirSwab
```


#Venn diagram and UpSet plot: overlapping taxa
```{r}
library(VennDiagram)
# Extract the taxa names from each object
taxa1 <- taxa_names(ps_ITS_sample)
taxa2 <- taxa_names(ITS_PosCon)
taxa3 <- taxa_names(ITS_ntc)
taxa5 <- taxa_names(ITS_nec)
taxa6 <- taxa_names(ITS_FB)
taxa7 <- taxa_names(ITS_MB)

# Combine into a list
taxa_list <- list(
  Sample = taxa1,
  PosControl = taxa2
)

# Create a Venn diagram
venn.plot <- venn.diagram(
  x = taxa_list,
  filename = NULL,        # Don't write to file immediately
  col = "black",
  fill = c("dodgerblue", "goldenrod1"),
  alpha = 0.50,
  cex = 1.5,
  cat.col = c("dodgerblue", "goldenrod1"),
  cat.cex = 1.5,
  cat.fontface = "bold",
  main = "Overlapping Taxa Among Four Phyloseq Objects"
)

grid::grid.draw(venn.plot)

# If you want to save it as an image file:
png(filename = "VennDiagram_ITS_taxa_wControl.png", width = 800, height = 800, res = 150)
grid::grid.draw(venn.plot)
dev.off()

#Investigate different types of control
taxa_list <- list(
  Sample = taxa1,
  PosControl = taxa2,
  ntc = taxa3,
  nec = taxa5,
  FieldBlank = taxa6,
  MethodBlank = taxa7
)

library("ggVennDiagram")
ggVennDiagram(taxa_list, force_upset = TRUE, order.set.by = "name", order.intersect.by = "none")
```

#Venn diagram and UpSet plot: overlapping taxa
```{r}
library(VennDiagram)
# Extract the taxa names from each object
taxa1 <- taxa_names(ps_ITS2_Sample)
taxa2 <- taxa_names(ps_ITS2_PosCon)
taxa3 <- taxa_names(ps_ITS2_ntc)
taxa4 <- taxa_names(ps_ITS2_nc)
taxa5 <- taxa_names(ps_ITS2_nec)
taxa6 <- taxa_names(ps_ITS2_FB)
taxa7 <- taxa_names(ps_ITS2_MB)

# Combine into a list
taxa_list <- list(
  Sample = taxa1,
  PosControl = taxa2
)

# Create a Venn diagram
venn.plot <- venn.diagram(
  x = taxa_list,
  filename = NULL,        # Don't write to file immediately
  col = "black",
  fill = c("dodgerblue", "goldenrod1"),
  alpha = 0.50,
  cex = 1.5,
  cat.col = c("dodgerblue", "goldenrod1"),
  cat.cex = 1.5,
  cat.fontface = "bold",
  main = "Overlapping Taxa Among Four Phyloseq Objects"
)

grid::grid.draw(venn.plot)

# If you want to save it as an image file:
png(filename = "VennDiagram_ITS2_taxa_wControl.png", width = 800, height = 800, res = 150)
grid::grid.draw(venn.plot)
dev.off()

#Investigate different types of control
taxa_list <- list(
  Sample = taxa1,
  PosControl = taxa2,
  ntc = taxa3,
  nc = taxa4,
  nec = taxa5,
  FieldBlank = taxa6,
  MethodBlank = taxa7
)

library("ggVennDiagram")
ggVennDiagram(taxa_list, force_upset = TRUE, order.set.by = "name", order.intersect.by = "none")
```

#Fig. S4 Abundance plot
```{r}
#Illumina ITS2
# Step 1: Aggregate taxa at desired rank, e.g., Phylum
ps_abundance <- tax_glom(ps_Illumina_ITS2_filt, taxrank = "phylum")

# Step 2: Convert to relative abundances
ps_abundance <- microbiome::transform(ps_abundance, "compositional")

# Step 3: Create a stacked bar plot
p_ITS2abundance <- plot_bar(ps_abundance, fill = "phylum") + 
     geom_bar(stat = "identity", position = "stack") +
     theme_void() +
     theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
     ggtitle("Relative Abundance of Taxa at Phylum Level - Short Read")

print(p_ITS2abundance)

#PacBio ITS
# Step 1: Aggregate taxa at desired rank, e.g., Phylum
ps_abundance <- tax_glom(ps_PacBio_ITS_filt, taxrank = "Phylum")

# Step 2: Convert to relative abundances
ps_abundance <- microbiome::transform(ps_abundance, "compositional")

# Step 3: Create a stacked bar plot
p_ITSabundance <- plot_bar(ps_abundance, fill = "Phylum") + 
     geom_bar(stat = "identity", position = "stack") +
     theme_void() +
     theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
     ggtitle("Relative Abundance of Taxa at Phylum Level - Long Read")

print(p_ITSabundance)

#PacBio ITS2
# Step 1: Aggregate taxa at desired rank, e.g., Phylum
ps_abundance <- tax_glom(ps_PacBio_ITS2_filt, taxrank = "Phylum")

# Step 2: Convert to relative abundances
ps_abundance <- microbiome::transform(ps_abundance, "compositional")

# Step 3: Create a stacked bar plot
p_ITS2abundance <- plot_bar(ps_abundance, fill = "Phylum") + 
     geom_bar(stat = "identity", position = "stack") +
     theme_void() +
     theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
     ggtitle("Relative Abundance of Taxa at Phylum Level - Long Read Trimmed")

print(p_ITS2abundance)
```

#Reads per sample
```{r}
# Get read counts per sample
reads_per_sample <- sample_sums(ps_Illumina_ITS2)

# Calculate mean and standard deviation
mean_reads <- mean(reads_per_sample)
sd_reads <- sd(reads_per_sample)

# Print results
cat("Mean reads per sample Illumina ITS2:", round(mean_reads, 2), "\n")
cat("Standard deviation Illumina ITS2:", round(sd_reads, 2), "\n")
ps_Illumina_ITS2

# Get read counts per sample
reads_per_sample <- sample_sums(ps_PacBio_Fungi_match)

# Calculate mean and standard deviation
mean_reads <- mean(reads_per_sample)
sd_reads <- sd(reads_per_sample)

# Print results
cat("Mean reads per sample PacBio ITS:", round(mean_reads, 2), "\n")
cat("Standard deviation PacBio ITS:", round(sd_reads, 2), "\n")
ps_PacBio_Fungi_match
```

#Internal QC: read and ASV distribution
```{r}
illum_reads <- sample_sums(ps_ITS2_Sample)
pacbio_reads <- sample_sums(ps_ITS_sample)

df_qc <- bind_rows(
  tibble(sample = names(illum_reads),
         reads = illum_reads,
         platform = "ITS2"),
  tibble(sample = names(pacbio_reads),
         reads = pacbio_reads,
         platform = "Full_ITS")
)

#Plot read per sample
ggplot(df_qc, aes(x = platform, y = reads, fill = platform)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  scale_fill_manual(values = c("ITS2"="#F28E2B","Full_ITS"="#B91372")) +
  labs(x = "", y = "Post-QC reads per sample") +
  theme_classic(base_size = 14)


illum_asvs <- apply(otu_table(ps_ITS2_Sample), 2, function(x) sum(x > 0))
pacbio_asvs <- apply(otu_table(ps_ITS_sample), 2, function(x) sum(x > 0))

df_asv <- bind_rows(
  tibble(sample = names(illum_asvs), asvs = illum_asvs, platform = "ITS2"),
  tibble(sample = names(pacbio_asvs), asvs = pacbio_asvs, platform = "Full_ITS")
)


ggplot(df_asv, aes(x = platform, y = asvs, fill = platform)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  labs(x = "", y = "ASVs per sample") +
  theme_classic(base_size = 14)
```


#ASV assigned
##Illumina
```{r}
tax <- data.frame(ps_Illumina_ITS2_filt@tax_table)
#Optional
#tax$Species[grepl("_sp$", tax$Species)] <- NA
#tax$Genus[grepl("_gen_Incertae_sedis$", tax$Genus)] <- NA
#tax$Gamily[grepl("_fam_Incertae_sedis$", tax$Family)] <- NA
#tax$Order[grepl("_ord_Incertae_sedis$", tax$Order)] <- NA
#tax$Class[grepl("_cls_Incertae_sedis$", tax$Class)] <- NA
#tax$Phylum[grepl("_phy_Incertae_sedis$", tax$Phylum)] <- NA
#tax$Kingdom[grepl("Unassigned", tax$Kingdom)] <- NA

# Count the number of ASVs assigned at each level
assigned_counts <- colSums(!is.na(tax))

# Calculate the total number of ASVs
total_asvs <- nrow(tax)

# Calculate the percentage of ASVs assigned at each level
assigned_percentages <- (assigned_counts / total_asvs) * 100

# Combine into a data frame for better visualization
assignment_stats <- data.frame(
  Taxonomy_Level = colnames(tax),
  Assigned_Count = assigned_counts,
  Total_ASVs = total_asvs,
  Assigned_Percentage = assigned_percentages
)

Illumina_assignment <- assignment_stats[1:7, ]

# Arrange the orders of taxonomy rank
taxonomy_order <- c("kingdom", "phylum", "class", "order", "family", "genus", "species")
Illumina_assignment$Taxonomy_Level <- factor(Illumina_assignment$Taxonomy_Level, levels = taxonomy_order)

# Print the results
print(Illumina_assignment)

#Plot
ASV_assigned_distribution <- ggplot(Illumina_assignment, aes(x = Taxonomy_Level, y = Assigned_Percentage)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.2, fill = "#7BAFD4", alpha = 0.7) +
  geom_text(aes(label = round(Assigned_Percentage, 1)),  # Add labels rounded to 1 decimal place
            vjust = -0.2,   # Position labels just above the bars
            color = "black", # Color of the text labels
            size = 4) +      # Adjust the size of the labels
  labs(title = "Percentage of ASVs Assigned at Each Taxonomy Level",
       x = "Taxonomy Level", y = "Assigned Percentage (%)") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )

ASV_assigned_distribution
```

##PacBio_Fungi
```{r}
tax_PacBio_fungi <- data.frame(ps_PacBio_ITS_filt@tax_table)
#Optional
#tax_PacBio_fungi$Species[grepl("_sp$", tax_PacBio_fungi$Species)] <- NA
#tax_PacBio_fungi$Genus[grepl("_gen_Incertae_sedis$", tax_PacBio_fungi$Genus)] <- NA
#tax_PacBio_fungi$Family[grepl("_fam_Incertae_sedis$", tax_PacBio_fungi$Family)] <- NA
#tax_PacBio_fungi$Order[grepl("_ord_Incertae_sedis$", tax_PacBio_fungi$Order)] <- NA
#tax_PacBio_fungi$Class[grepl("_cls_Incertae_sedis$", tax_PacBio_fungi$Class)] <- NA
#tax_PacBio_fungi$Phylum[grepl("_phy_Incertae_sedis$", tax_PacBio_fungi$Phylum)] <- NA

# Count the number of ASVs assigned at each level
assigned_counts <- colSums(!is.na(tax_PacBio_fungi))

# Calculate the total number of ASVs
total_asvs <- nrow(tax_PacBio_fungi)

# Calculate the percentage of ASVs assigned at each level
assigned_percentages <- (assigned_counts / total_asvs) * 100

# Combine into a data frame for better visualization
assignment_stats <- data.frame(
  Taxonomy_Level = colnames(tax_PacBio_fungi),
  Assigned_Count = assigned_counts,
  Total_ASVs = total_asvs,
  Assigned_Percentage = assigned_percentages
)

PacBio_assignment <- assignment_stats[1:7, ]

# Arrange the orders of taxonomy rank
taxonomy_order <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
PacBio_assignment$Taxonomy_Level <- factor(PacBio_assignment$Taxonomy_Level, levels = taxonomy_order)

# Print the results
print(PacBio_assignment)

#Plot
ASV_assigned_distribution <- ggplot(PacBio_assignment, aes(x = Taxonomy_Level, y = Assigned_Percentage)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.2, fill = "#7BAFD4", alpha = 0.7) +
  geom_text(aes(label = round(Assigned_Percentage, 1)),  # Add labels rounded to 1 decimal place
            vjust = -0.2,   # Position labels just above the bars
            color = "black", # Color of the text labels
            size = 4) +      # Adjust the size of the labels
  labs(title = "Percentage of ASVs Assigned at Each Taxonomy Level",
       x = "Taxonomy Level", y = "Assigned Percentage (%)") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )

ASV_assigned_distribution
```

##PacBio_ITS2
```{r}
tax_pacbio_ITS2 <- data.frame(ps_PacBio_ITS2_filt@tax_table)
#Optional
tax_pacbio_ITS2$Species[grepl("_sp$", tax_pacbio_ITS2$Species)] <- NA
tax_pacbio_ITS2$Genus[grepl("_gen_Incertae_sedis$", tax_pacbio_ITS2$Genus)] <- NA
tax_pacbio_ITS2$Family[grepl("_fam_Incertae_sedis$", tax_pacbio_ITS2$Family)] <- NA
tax_pacbio_ITS2$Order[grepl("_ord_Incertae_sedis$", tax_pacbio_ITS2$Order)] <- NA
tax_pacbio_ITS2$Class[grepl("_cls_Incertae_sedis$", tax_pacbio_ITS2$Class)] <- NA
tax_pacbio_ITS2$Phylum[grepl("_phy_Incertae_sedis$", tax_pacbio_ITS2$Phylum)] <- NA

# Count the number of ASVs assigned at each level
assigned_counts <- colSums(!is.na(tax_pacbio_ITS2))

# Calculate the total number of ASVs
total_asvs <- nrow(tax_pacbio_ITS2)

# Calculate the percentage of ASVs assigned at each level
assigned_percentages <- (assigned_counts / total_asvs) * 100

# Combine into a data frame for better visualization
assignment_stats <- data.frame(
  Taxonomy_Level = colnames(tax_pacbio_ITS2),
  Assigned_Count = assigned_counts,
  Total_ASVs = total_asvs,
  Assigned_Percentage = assigned_percentages
)

PacBio_ITS2_assignment <- assignment_stats[1:7, ]

# Arrange the orders of taxonomy rank
taxonomy_order <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
PacBio_ITS2_assignment$Taxonomy_Level <- factor(PacBio_assignment$Taxonomy_Level, levels = taxonomy_order)

# Print the results
print(PacBio_ITS2_assignment)

#Plot
ASV_assigned_distribution <- ggplot(PacBio_ITS2_assignment, aes(x = Taxonomy_Level, y = Assigned_Percentage)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.2, fill = "deeppink", alpha = 0.7) +
  geom_text(aes(label = round(Assigned_Percentage, 1)),  # Add labels rounded to 1 decimal place
            vjust = -0.2,   # Position labels just above the bars
            color = "black", # Color of the text labels
            size = 4) +      # Adjust the size of the labels
  labs(title = "Percentage of ASVs Assigned at Each Taxonomy Level",
       x = "Taxonomy Level", y = "Assigned Percentage (%)") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )

ASV_assigned_distribution
```


#%Assigned compare
```{r}
PacBio_assignment$Method <- "Full-length ITS"
PacBio_ITS2_assignment$Method <- "Trimmed ITS2"
Illumina_assignment$Method <- "ITS2"
Illumina_assignment <- Illumina_assignment %>%
  mutate(Taxonomy_Level = str_to_title(Taxonomy_Level))

combined_assignment <- bind_rows(PacBio_assignment, Illumina_assignment, PacBio_ITS2_assignment)

# Arrange the orders of taxonomy rank
taxonomy_order <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
combined_assignment$Taxonomy_Level <- factor(combined_assignment$Taxonomy_Level, levels = taxonomy_order)

PercentAssigned_combined <- ggplot(combined_assignment, aes(x = Taxonomy_Level, y = Assigned_Percentage, fill = Method)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.2, position = "dodge") +
  geom_text(aes(label = round(Assigned_Percentage, 1)),  # Add labels rounded to 1 decimal place
            vjust = -0.2,   # Position labels just above the bars
            color = "black", # Color of the text labels
            size = 4) + 
  labs(title = "Percentage of ASVs Assigned",
       x = "Taxonomic Rank", y = "% Assigned") +
  scale_fill_manual(values = c("Full-length ITS" = "deeppink", "ITS2" = "orange","Trimmed ITS2" = "pink")) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )

PercentAssigned_combined
```

# Fig 1a: %Assigned paired comparison by sample
```{r}
percent_assigned_by_Rank <- function(ps, ranks = c("Kingdom","Phylum","Class","Order","Family","Genus","Species")) {
  stopifnot(inherits(ps, "phyloseq"))

  OTU  <- as(otu_table(ps), "matrix")        # taxa x samples (or samples x taxa)
  if (taxa_are_rows(ps)) OTU <- t(OTU)       # make it samples x taxa
  TAX  <- as(tax_table(ps), "matrix")        # taxa x ranks

  # Ensure rank names exist
  ranks <- ranks[ranks %in% colnames(TAX)]
  if (length(ranks) == 0) stop("None of the specified ranks found in tax_table(ps).")

  # sample totals (denominator)
  sample_totals <- rowSums(OTU)
  keep_samples  <- sample_totals > 0
  OTU <- OTU[keep_samples, , drop = FALSE]
  sample_totals <- sample_totals[keep_samples]

  # For each rank, mark taxa that are assigned (non‑NA and non‑empty)
  is_assigned_by_rank <- lapply(ranks, function(rk) {
    tax_vec <- TAX[, rk]
    !is.na(tax_vec) & nzchar(tax_vec)
  })
  names(is_assigned_by_rank) <- ranks

  # Compute % assigned per sample per rank
  df_list <- lapply(ranks, function(rk) {
    assigned_taxa <- is_assigned_by_rank[[rk]]
    # abundance‑weighted numerator
    num <- OTU[, assigned_taxa, drop = FALSE] |> rowSums()
    pct <- 100 * num / sample_totals

    # # If you prefer unweighted (by ASV presence), replace the 2 lines above by:
    # present <- OTU > 0
    # pct <- 100 * rowSums(present[, assigned_taxa, drop = FALSE]) / rowSums(present)

    tibble(sample_id = rownames(OTU),
           rank      = rk,
           pct_assigned = as.numeric(pct))
  })

  bind_rows(df_list)
}

percent_assigned_by_rank <- function(ps, ranks = c("kingdom","phylum","class","order","family","genus","species")) {
  stopifnot(inherits(ps, "phyloseq"))

  OTU  <- as(otu_table(ps), "matrix")        # taxa x samples (or samples x taxa)
  if (taxa_are_rows(ps)) OTU <- t(OTU)       # make it samples x taxa
  TAX  <- as(tax_table(ps), "matrix")        # taxa x ranks

  # Ensure rank names exist
  ranks <- ranks[ranks %in% colnames(TAX)]
  if (length(ranks) == 0) stop("None of the specified ranks found in tax_table(ps).")

  # sample totals (denominator)
  sample_totals <- rowSums(OTU)
  keep_samples  <- sample_totals > 0
  OTU <- OTU[keep_samples, , drop = FALSE]
  sample_totals <- sample_totals[keep_samples]

  # For each rank, mark taxa that are assigned (non‑NA and non‑empty)
  is_assigned_by_rank <- lapply(ranks, function(rk) {
    tax_vec <- TAX[, rk]
    !is.na(tax_vec) & nzchar(tax_vec)
  })
  names(is_assigned_by_rank) <- ranks

  # Compute % assigned per sample per rank
  df_list <- lapply(ranks, function(rk) {
    assigned_taxa <- is_assigned_by_rank[[rk]]
    # abundance‑weighted numerator
    num <- OTU[, assigned_taxa, drop = FALSE] |> rowSums()
    pct <- 100 * num / sample_totals

    # # If you prefer unweighted (by ASV presence), replace the 2 lines above by:
    # present <- OTU > 0
    # pct <- 100 * rowSums(present[, assigned_taxa, drop = FALSE]) / rowSums(present)

    tibble(sample_id = rownames(OTU),
           rank      = rk,
           pct_assigned = as.numeric(pct))
  })

  bind_rows(df_list)
}
```


```{r}
# Choose the ranks you want to test
Ranks <- c("Kingdom","Phylum","Class","Order","Family","Genus","Species")
ranks <- c("kingdom","phylum","class","order","family","genus","species")

#Compute %assigned for each platform and combine
ITS2_df <- percent_assigned_by_rank(ps_Illumina_ITS2_filt, ranks) %>%
  mutate(platform = "ITS2") %>%
  mutate(rank = str_to_title(rank))

ITS_df <- percent_assigned_by_Rank(ps_PacBio_ITS_filt, Ranks) %>%
  mutate(platform = "Full_ITS")

#Make a paired wide table by rank
pct_wide <- ITS2_df %>%
  select(sample_id, rank, pct_assigned) %>%
  rename(ITS2 = pct_assigned) %>%
  inner_join(
    ITS_df %>%
      select(sample_id, rank, pct_assigned) %>%
      rename(Full_ITS = pct_assigned),
    by = c("sample_id", "rank")
  ) %>%
  mutate(diff_pct = Full_ITS - ITS2)

pct_wide %>% count(rank)

#Run paired Wilcoxon per rank
run_wilcox <- function(dat) {
  wt <- wilcox.test(dat$Full_ITS, dat$ITS2,
                    paired = TRUE, alternative = "two.sided",
                    conf.int = TRUE, conf.level = 0.95,
                    exact = FALSE, correct = FALSE)
  tibble::tibble(
    n_pairs     = nrow(dat),
    V           = unname(wt$statistic),
    p_value     = wt$p.value,
    median_diff = unname(wt$estimate),  # HL estimate of median(FullITS − ITS2)
    conf_low    = wt$conf.int[1],
    conf_high   = wt$conf.int[2]
  )
}

wilcox_by_rank <- pct_wide %>%
  group_by(rank) %>%
  group_modify(~ run_wilcox(.x)) %>%
  ungroup() %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

wilcox_by_rank
```

```{r}
# Plot
pct_wide_long <- pct_wide %>%
  pivot_longer(cols = c(ITS2, Full_ITS),
               names_to = "platform",
               values_to = "pct_assigned")

sum_df <- pct_wide_long %>%
  group_by(rank, platform) %>%
  summarise(
    mean_pct = mean(pct_assigned, na.rm = TRUE),
    sd_pct   = sd(pct_assigned, na.rm = TRUE),
    se_pct   = sd_pct / sqrt(n()),
    ci_low   = mean_pct - qt(0.975, df = n()-1) * se_pct,
    ci_up    = mean_pct + qt(0.975, df = n()-1) * se_pct,
    n = n(),
    .groups = "drop"
  )

taxonomy_order <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
sum_df$rank <- factor(sum_df$rank, levels = taxonomy_order)

PCT_Assigned <- ggplot(sum_df, aes(x = rank, y = mean_pct, fill = platform)) +
  geom_col(position = position_dodge(width = 0.7),
           width = 0.6, color = "black", linewidth = 0.5) +
  geom_text(aes(label = round(mean_pct, 1)),  # Add labels rounded to 1 decimal place
            vjust = -1.0,   # Position labels just above the bars
            color = "black", # Color of the text labels
            position = position_dodge(width = 0.7),           
             size = 4) + 
  geom_errorbar(aes(ymin = mean_pct - se_pct, ymax = mean_pct + se_pct),
                position = position_dodge(width = 0.7),
                width = 0.25, linewidth = 0.5) +
  scale_fill_manual(values = c(Full_ITS = "deeppink", ITS2 = "orange"),
                    labels = c("Full-length ITS", "ITS2")) +
  labs(x = "Taxonomic Rank",
       y = "% Assigned",
       fill = "Method") +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

PCT_Assigned


diff_tbl <- pct_wide %>%
  filter(rank != "Kingdom") %>%
  mutate(diff_pct = Full_ITS - ITS2)

diff_sum <- diff_tbl %>%
  group_by(rank) %>%
  summarise(
    mean_d = mean(diff_pct),
    se_d   = sd(diff_pct) / sqrt(n()),
    ci95_l = mean_d - qt(0.975, df = n()-1) * se_d,
    ci95_u = mean_d + qt(0.975, df = n()-1) * se_d,
    .groups = "drop"
  )

taxonomy_order <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
diff_sum$rank <- factor(diff_sum$rank, levels = taxonomy_order)

diff_assigned <- ggplot(diff_sum, aes(x = rank, y = mean_d)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(fill = "#6C83B5", color = "grey20", width = 0.6) +
  geom_errorbar(aes(ymin = ci95_l, ymax = ci95_u), width = 0.2) +
  labs(x = "Taxonomic rank", y = "Full‑length ITS − ITS2 (% assigned)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

diff_assigned
```

# Fig S7: PacBio trimmed ITS2 assignment by sample
```{r}
Trimmed_ITS2_df <- percent_assigned_by_Rank(ps_PacBio_ITS2_filt, Ranks) %>%
  mutate(platform = "Trimmed_ITS2")

sum_df_ITS2 <- Trimmed_ITS2_df %>%
  group_by(rank) %>%
  summarise(
    mean_pct = mean(pct_assigned, na.rm = TRUE),
    sd_pct   = sd(pct_assigned, na.rm = TRUE),
    se_pct   = sd_pct / sqrt(n()),
    ci_low   = mean_pct - qt(0.975, df = n()-1) * se_pct,
    ci_up    = mean_pct + qt(0.975, df = n()-1) * se_pct,
    n = n(),
    .groups = "drop"
  )

taxonomy_order <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
sum_df_ITS2$rank <- factor(sum_df_ITS2$rank, levels = taxonomy_order)

PCT_Assigned <- ggplot(sum_df_ITS2, aes(x = rank, y = mean_pct)) +
  geom_col(position = position_dodge(width = 0.7),
           width = 0.6, color = "black", linewidth = 0.5) +
  geom_text(aes(label = round(mean_pct, 1)),  # Add labels rounded to 1 decimal place
            vjust = -1.0,   # Position labels just above the bars
            color = "black", # Color of the text labels
            position = position_dodge(width = 0.7),           
             size = 4) + 
  geom_errorbar(aes(ymin = mean_pct - se_pct, ymax = mean_pct + se_pct),
                position = position_dodge(width = 0.7),
                width = 0.25, linewidth = 0.5) +
  geom_col(fill = "pink", color = "grey20", width = 0.6) +
  labs(x = "Taxonomic Rank",
       y = "% Assigned") +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

PCT_Assigned
```


# Fig S3: Sequence length distribution
```{r}
ps_Illumina_ITS2

sequence_lengths <- width(ps_Illumina_ITS2@refseq)
length_df <- data.frame(Length = sequence_lengths)

# Histogram of sequence lengths
sequence_length_distribution <- ggplot(length_df, aes(x = Length)) +
  geom_histogram(binwidth = 10, color = "black", linewidth = 0.2, fill = "orange", alpha = 0.7) +
  labs(title = "Sequence Length Distribution", x = "Sequence Length (bp)", y = "Frequency") +
  theme_bw()+
   theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank()  # Remove minor grid lines
  )

sequence_length_distribution
```

```{r}
ps_PacBio_ITS2_match

sequence_lengths <- width(ps_PacBio_ITS2_match@refseq)
length_df <- data.frame(Length = sequence_lengths)

# Histogram of sequence lengths
sequence_length_distribution <- ggplot(length_df, aes(x = Length)) +
  geom_histogram(binwidth = 10, color = "black", linewidth = 0.2, fill = "pink", alpha = 0.7) +
  labs(title = "Sequence Length Distribution", x = "Sequence Length (bp)", y = "Frequency") +
  theme_bw()+
   theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank()  # Remove minor grid lines
  )

sequence_length_distribution
```

```{r}
ps_PacBio_Fungi_match

sequence_lengths <- width(ps_PacBio_Fungi_match@refseq)
length_df <- data.frame(Length = sequence_lengths)

# Histogram of sequence lengths
sequence_length_distribution <- ggplot(length_df, aes(x = Length)) +
  geom_histogram(binwidth = 10, color = "black", linewidth = 0.2, fill = "deeppink", alpha = 0.7) +
  labs(title = "Sequence Length Distribution", x = "Sequence Length (bp)", y = "Frequency") +
  theme_bw()+
   theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank()  # Remove minor grid lines
  )

sequence_length_distribution
```

#Fig S8: Ordination plot with ggplot
```{r}
ps_Illumina_ITS2_filt
ps_PacBio_ITS_filt
ps_trimmed_ITS2_merged

#Transform to relative abundance
ps_Illumina_ITS2_rel <- microbiome::transform(ps_Illumina_ITS2_filt,"compositional")
ps_PacBio_Fungi_match_rel <- microbiome::transform(ps_PacBio_ITS_filt,"compositional")
ps_PacBio_ITS2_rel <- microbiome::transform(ps_trimmed_ITS2_merged,"compositional")

#Make sure sample name match
common_samples <- intersect(sample_names(ps_Illumina_ITS2_rel), 
                            sample_names(ps_PacBio_Fungi_match_rel))

ps_Illumina_common <- prune_samples(common_samples, ps_Illumina_ITS2_rel)
ps_PacBio_common   <- prune_samples(common_samples, ps_PacBio_Fungi_match_rel)
ps_PacBio_ITS2_common <- prune_samples(common_samples, ps_PacBio_ITS2_rel)

ps_Illumina_common <- prune_taxa(taxa_sums(ps_Illumina_common) > 0, ps_Illumina_common)
ps_PacBio_common   <- prune_taxa(taxa_sums(ps_PacBio_common) > 0, ps_PacBio_common)
ps_PacBio_ITS2_common <- prune_taxa(taxa_sums(ps_PacBio_ITS2_common) > 0, ps_PacBio_ITS2_common)

#PCoA with bray-cutis
ord_ITS <- ordinate(ps_PacBio_common, method="PCoA", distance = "bray")
p_ord_ITS <- plot_ordination(ps_PacBio_common, ord_ITS, color = "Location", shape = "SampleType") + 
  geom_point(size = 4, alpha = 0.8) +
  stat_ellipse(aes(group = Location), type = "norm", linetype = "dashed", size = 1, alpha = 0.5) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PCoA of Full-length ITS",
    subtitle = paste0("Bray-Curtis distance | Axis 1: ", round(ord_ITS$values$Relative_eig[1]*100, 1), 
                      "%, Axis 2: ", round(ord_ITS$values$Relative_eig[2]*100, 1), "%"),
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12)
  )
p_ord_ITS

ord_trimmedITS2 <- ordinate(ps_PacBio_ITS2_common, method="PCoA", distance = "bray")
p_ord_trimmedITS <- plot_ordination(ps_PacBio_ITS2_common, ord_trimmedITS2, color = "Location", shape = "SampleType") + 
  geom_point(size = 4, alpha = 0.8) +
  stat_ellipse(aes(group = Location), type = "norm", linetype = "dashed", size = 1, alpha = 0.5) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PCoA of Trimmed ITS2",
    subtitle = paste0("Bray-Curtis distance | Axis 1: ", round(ord_trimmedITS2$values$Relative_eig[1]*100, 1), 
                      "%, Axis 2: ", round(ord_trimmedITS2$values$Relative_eig[2]*100, 1), "%"),
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12)
  )
p_ord_trimmedITS

ord_ITS2 <- ordinate(ps_Illumina_common, method="PCoA", distance = "bray")
p_ord_ITS2 <- plot_ordination(ps_Illumina_common, ord_ITS2, color = "Location", shape = "SampleType") + 
  geom_point(size = 4, alpha = 0.8) +
  stat_ellipse(aes(group = Location), type = "norm", linetype = "dashed", size = 1, alpha = 0.5) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "PCoA of ITS2",
    subtitle = paste0("Bray-Curtis distance | Axis 1: ", round(ord_ITS2$values$Relative_eig[1]*100, 1), 
                      "%, Axis 2: ", round(ord_ITS2$values$Relative_eig[2]*100, 1), "%"),
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12)
  )
p_ord_ITS2

library(gridExtra)
grid.arrange(p_ord_ITS, p_ord_ITS2, ncol = 2)
```

##PERMANOVA
```{r}
library(vegan)

#Illumina ITS2
#transform the phyloseq if not yet
ps <- ps_Illumina_common

df <- t(data.frame(ps@otu_table))
metadata <- data.frame(ps@sam_data)

#PERMANOVA
permanova <- adonis2(df ~ SampleType * Location, data = metadata, method = "bray", by = "terms", permutations = 10000)
print(permanova)

#PacBio full-length ITS
#transform the phyloseq if not yet
ps <- ps_PacBio_common

df <- t(data.frame(ps@otu_table))
metadata <- data.frame(ps@sam_data)

#PERMANOVA
permanova <- adonis2(df ~ SampleType * Location, data = metadata, method = "bray", by = "terms", permutations = 10000)
print(permanova)

#Trimmed ITS2
#transform the phyloseq if not yet
ps <- ps_PacBio_ITS2_common

df <- t(data.frame(ps@otu_table))
metadata <- data.frame(ps@sam_data)

#PERMANOVA
permanova <- adonis2(df ~ SampleType * Location, data = metadata, method = "bray", by = "terms", permutations = 10000)
print(permanova)
```

##Mental test and procrustes analysis
```{r}
# Compute Bray–Curtis distances
dist_ITS2 <- phyloseq::distance(ps_Illumina_common, method = "bray")
dist_ITS <- phyloseq::distance(ps_PacBio_common, method = "bray")
dist_trimmedITS2 <- phyloseq::distance(ps_PacBio_ITS2_common, method = "bray")

# Mantel test
mantel_result <- mantel(dist_ITS2, dist_ITS, method = "spearman", permutations = 999)
print(mantel_result)

mantel_result <- mantel(dist_ITS2, dist_trimmedITS2, method = "spearman", permutations = 999)
print(mantel_result)

# Procrustes analysis
## Extract coordinates
coords_ITS2 <- ord_ITS2$vectors
coords_ITS <- ord_ITS$vectors
coords_trimmedITS2 <- ord_trimmedITS2$vectors

## Procrustes analysis: ITS2 vs full-length ITS
proc <- procrustes(coords_ITS, coords_ITS2, scale = TRUE)
proc_test <- protest(coords_ITS, coords_ITS2, permutations = 999)
summary(proc_test)

proc_test$t0
proc_test$signif

## Procrustes analysis: ITS2 vs trimmed ITS2
proc <- procrustes(coords_trimmedITS2, coords_ITS2, scale = TRUE)
proc_test <- protest(coords_trimmedITS2, coords_ITS2, permutations = 999)
summary(proc_test)

proc_test$t0
proc_test$signif
proc_ss  <- 1 - (proc_test$t0)^2 
proc_ss

# Make the table
mantel_r <- mantel_result$statistic
mantel_p <- mantel_result$signif
proc_r   <- proc_test$t0        # 0.6663367
proc_p   <- proc_test$signif    # 0.002
proc_ss  <- 1 - proc_r^2        # 0.5559954

inset_tbl <- data.frame(
  Metric = c("Mantel (Spearman) on Bray–Curtis", "Procrustes (symmetric)"),
  Statistic = c(sprintf("r = %.3f", mantel_r),
                sprintf("r = %.3f (ss = %.3f)", proc_r, proc_ss)),
  P_value = c(sprintf("%.3f", mantel_p),
              sprintf("%.3f", proc_p)),
  stringsAsFactors = FALSE
)

inset_tbl

#Robustness check
k <- 15
coords_ITS2_k <- ord_ITS2$vectors[, 1:k, drop = FALSE]
coords_ITS_k <- ord_ITS$vectors[, 1:k, drop = FALSE]
proc_test_k <- protest(coords_ITS_k, coords_ITS2_k, permutations = 999)
c(r = proc_test_k$t0, p = proc_test_k$signif)
```

##Fig 2b: Paired Ordination and PERMANOVA
```{r}
#Prepare data
sample_data(ps_Illumina_common)$Workflow <- "ITS2"
sample_data(ps_PacBio_common)$Workflow  <- "Full-length ITS"

sample_names(ps_Illumina_common) <- paste0(sample_names(ps_Illumina_common), "_Illumina")
sample_names(ps_PacBio_common)   <- paste0(sample_names(ps_PacBio_common), "_PacBio")

sample_data(ps_Illumina_common)$PairID <- gsub("_Illumina$", "", sample_names(ps_Illumina_common))
sample_data(ps_PacBio_common)$PairID   <- gsub("_PacBio$", "", sample_names(ps_PacBio_common))

taxa_names(ps_Illumina_common) <- paste0("ITS2_", taxa_names(ps_Illumina_common))
taxa_names(ps_PacBio_common)   <- paste0("ITSfull_", taxa_names(ps_PacBio_common))

ps_combined <- merge_phyloseq(ps_Illumina_common, ps_PacBio_common)
ps_combined <- prune_samples(sample_sums(ps_combined) > 0, ps_combined)

#Bray–Curtis distance
dist.bc <- phyloseq::distance(ps_combined, method = "bray")

adonis2(
  dist.bc ~ Workflow + SampleType + Location,
  data = data.frame(sample_data(ps_combined)),
  permutations = 10000,
  strata = sample_data(ps_combined)$PairID,
  by = "terms"
)

#Plot
ord <- ordinate(ps_combined, method = "PCoA", distance = "bray")
paired_ordination <- plot_ordination(ps_combined, ord, shape = "Workflow", color = "Location") + 
  geom_point(size = 2.5, alpha = 0.8) +
  facet_wrap(~ SampleType) + 
  stat_ellipse(aes(group = Workflow), type = "norm", linetype = "dashed", size = 1, alpha = 0.5) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_manual(values = c(
  "Full-length ITS" = "deeppink",
  "ITS2" = "orange"
)) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12)
  )
paired_ordination + geom_line(aes(group = PairID),color = "grey60", alpha = 0.4)

paired_ordination <- plot_ordination(ps_combined, ord, color = "Workflow", shape = "SampleType") + 
   geom_point(size = 2.5, alpha = 0.8) +
   facet_wrap(~ Location) + 
   stat_ellipse(aes(group = Workflow), type = "norm", linetype = "dashed", size = 1, alpha = 0.5) +
   scale_color_manual(values = c(
   "Full-length ITS" = "deeppink",
   "ITS2" = "orange")) + 
   scale_fill_manual(values = c(
   "Full-length ITS" = "deeppink",
   "ITS2" = "orange")) +
   theme_bw(base_size = 14) +
   theme(
     panel.grid = element_blank(),
     legend.position = "right",
     plot.title = element_text(face = "bold", size = 18),
     plot.subtitle = element_text(size = 12)
   )

paired_ordination + geom_line(aes(group = PairID),color = "grey60", alpha = 0.4)
```

##Fig S6: Scatter PCoA
```{r}
# Run PCoA
pcoa_res <- ordinate(ps_combined, method = "PCoA", distance = dist.bc)

# Extract sample scores
pcoa_df <- as.data.frame(pcoa_res$vectors)
pcoa_df$SampleID <- rownames(pcoa_df)

# Add metadata
meta_df <- data.frame(sample_data(ps_combined))
meta_df$SampleID <- rownames(meta_df)

pcoa_df <- left_join(pcoa_df, meta_df, by = "SampleID")

# Keep only the first two axes
pcoa_long <- pcoa_df %>%
  select(PairID, Workflow, Axis.1, Axis.2) %>%
  pivot_wider(
    id_cols = PairID,
    names_from = Workflow,
    values_from = c(Axis.1, Axis.2)
  )

# Pearson correlation for PC1
cor_pc1 <- cor(pcoa_long$Axis.1_ITS2,
               pcoa_long$`Axis.1_Full-length ITS`,
               method = "pearson",
               use = "complete.obs")

r2_pc1 <- cor_pc1^2

ggplot(pcoa_long,
       aes(x = Axis.1_ITS2, y = `Axis.1_Full-length ITS`)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey40") +
  annotate("text",
           x = min(pcoa_long$Axis.1_ITS2, na.rm = TRUE),
           y = max(pcoa_long$`Axis.1_Full-length ITS`, na.rm = TRUE),
           hjust = 0, vjust = 1,
           label = paste0("R² = ", round(r2_pc1, 2),
                          "\nr = ", round(cor_pc1, 2))) +
  labs(x = "PCoA Axis 1 (ITS2)",
       y = "PCoA Axis 1 (Full-length ITS)",
       title = "Paired PCoA Axis 1 comparison") +
  theme_classic(base_size = 14)

# Pearson correlation for PC2
cor_pc2 <- cor(pcoa_long$Axis.2_ITS2,
               pcoa_long$`Axis.2_Full-length ITS`,
               method = "pearson",
               use = "complete.obs")

r2_pc2 <- cor_pc2^2

ggplot(pcoa_long,
       aes(x = Axis.2_ITS2, y = `Axis.2_Full-length ITS`)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey40") +
  annotate("text",
           x = min(pcoa_long$Axis.2_ITS2, na.rm = TRUE),
           y = max(pcoa_long$`Axis.2_Full-length ITS`, na.rm = TRUE),
           hjust = 0, vjust = 1,
           label = paste0("R² = ", round(r2_pc2, 2),
                          "\nr = ", round(cor_pc2, 2))) +
  labs(x = "PCoA Axis 2 (ITS2)",
       y = "PCoA Axis 2 (Full-length ITS)",
       title = "Paired PCoA Axis 2 comparison") +
  theme_classic(base_size = 14)

```


#Fig 2b: Scatter compare taxa abundance: ps_Illumina_ITS2 vs. ps_PacBio_Fungi_match do they agree
##Function: Log relative abundance
```{r}
log_relabund_transform <- function(ps, rank, pseudo_count = 1e-4) {
  library(tidyverse)
  counts <- otu_table(ps)
  if (taxa_are_rows(ps)) counts <- t(counts)
  rel_abund <- sweep(counts, 1, rowSums(counts), "/")
  log_relabund <- log10(rel_abund + pseudo_count)
  
  df <- as.data.frame(log_relabund)
  df$SampleID <- rownames(df)
  df_long <- pivot_longer(df, -SampleID, names_to = "FeatureID", values_to = "logRelAbund")
  
  tax <- as.data.frame(tax_table(ps))
  tax$FeatureID <- rownames(tax)

  if (!(rank %in% colnames(tax))) {
    stop(paste("Taxonomic rank", rank, "not found in taxonomy table."))
  }

  df_long <- left_join(df_long, tax[, c("FeatureID", rank)], by = "FeatureID") %>%
    dplyr::rename(Taxon = !!sym(rank)) %>%
    filter(!is.na(Taxon))

  return(df_long)
}
```

##Scatter plot
```{r}
ps_Illumina_ITS2_glom <- tax_glom(ps_Illumina_ITS2_filt, taxrank = "genus", NArm=TRUE, bad_empty=c(NA, "", " ", "\t"))
tax_ITS2_glom <- data.frame(ps_Illumina_ITS2_glom@tax_table)

ps_PacBio_ITS_match_glom <- tax_glom(ps_PacBio_ITS_filt, taxrank = "Genus", NArm=TRUE, bad_empty=c(NA, "", " ", "\t"))
tax_ITS_glom <- data.frame(ps_PacBio_ITS_match_glom@tax_table)

# For each taxa rank
rank_ITS <- c("Phylum", "Class", "Order", "Family", "Genus","Species")
rank_ITS2 <- c("phylum", "class", "order", "family", "genus","species")

# Store merged data per rank
merged_list <- list()

for (i in seq_along(rank_ITS)) {
  logrel_PacBio <- log_relabund_transform(ps_PacBio_ITS_match_glom, rank = rank_ITS[i])
  logrel_Illumina <- log_relabund_transform(ps_Illumina_ITS2_glom, rank = rank_ITS2[i])
  
  merged <- full_join(logrel_PacBio, logrel_Illumina, by = c("SampleID", "Taxon"), suffix = c("_PacBio", "_Illumina"))
  
  # NEW: Keep only rows where both values are not NA
  merged <- merged %>% filter(!is.na(logRelAbund_PacBio) & !is.na(logRelAbund_Illumina))
  
  merged$Rank <- rank_ITS[i]
  merged_list[[rank_ITS[i]]] <- merged
}

# Combine all ranks into one data frame
merged_all <- bind_rows(merged_list)

plot_rank_with_margins <- function(df, rank_label) {
  p <- ggplot(df, aes(x = logRelAbund_PacBio, y = logRelAbund_Illumina)) +
    geom_point(alpha = 0.4, color = "#2c3e50") +
    geom_smooth(method = "lm", se = FALSE, color = "#e74c3c") +
    stat_cor(method = "spearman", 
             aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
             label.x = min(df$logRelAbund_PacBio, na.rm = TRUE), 
             label.y = max(df$logRelAbund_Illumina, na.rm = TRUE)) +
    theme_classic(base_size = 14) +
    labs(title = paste("Long vs Short Read Log Relative Abundance at", rank_label),
         x = "Long read", y = "Short read") +
  theme(panel.border = element_rect(color = "black", fill = NA), # Adds an outer frame
        axis.line = element_line(color = "black"),               # Keeps axis lines
        panel.grid = element_blank())                            # Ensures no inner grid lines
  
  # Add marginal histograms
  ggMarginal(p, type = "density", fill = "gray70", alpha = 0.5)
}

```

##Fig 3b: P-value scatter by Location: from merged_list
```{r}
#At genus level match
overlapping_genus <- merged_list$Genus

#Add location
overlapping_genus <- overlapping_genus %>%
  left_join(
    metadata_ITS %>% select(Label, Location),
    by = c("SampleID" = "Label")
  )

table(overlapping_genus$Location)

library(broom)   # For tidy model results
library(purrr)

# Linear model
# Function to run per taxon and method
run_differential_abundance <- function(df, method_col) {
  df %>%
    group_by(Taxon) %>%
    nest() %>%
    mutate(
      model = map(data, ~ lm(reformulate("Location", method_col), data = .x)),
      tidied = map(model, tidy)
    ) %>%
    unnest(tidied) %>%
    filter(term != "(Intercept)") %>%
    select(Taxon, term, estimate, p.value)
}

#Run model for both pipelines
res_pacbio <- run_differential_abundance(overlapping_genus, "logRelAbund_PacBio") %>%
  mutate(
    fdr_pacbio = p.adjust(p.value, method = "BH"),
    signed_log10p_pacbio = sign(estimate) * -log10(p.value)
  )

res_illumina <- run_differential_abundance(overlapping_genus, "logRelAbund_Illumina") %>%
  mutate(
    fdr_illumina = p.adjust(p.value, method = "BH"),
    signed_log10p_illumina = sign(estimate) * -log10(p.value)
  )


#Merge results
comparison_df <- res_pacbio %>%
  rename(
    estimate_pacbio = estimate,
    pval_pacbio = p.value
  ) %>%
  inner_join(
    res_illumina %>%
      rename(
        estimate_illumina = estimate,
        pval_illumina = p.value
      ),
    by = c("Taxon", "term")
  ) %>%
  mutate(
    significant = ifelse(
      fdr_pacbio < 0.05 | fdr_illumina < 0.05,
      "FDR < 0.05",
      "n.s."
    )
  )

#Make the P-value vs P-value Plot
p_agreement_eachLocation <- ggplot(comparison_df,
       aes(x = estimate_illumina,
           y = estimate_pacbio,
           color = significant)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_text(
    data = comparison_df %>% filter(significant == "FDR < 0.05"),
    aes(label = Taxon),
    size = 3, hjust = 0, vjust = 1
  ) +
  #facet_wrap(~ term, scales = "free") +
  scale_color_manual(values = c("FDR < 0.05" = "firebrick", "n.s." = "gray60")) +
  labs(
    x = "Effect size (ITS2 – Illumina)",
    y = "Effect size (Full-length ITS – PacBio)",
    color = "Significance"
  ) +
  theme_classic(base_size = 14)

p_agreement_eachLocation 

# p-value vs p-value plot
library(ggrepel)  # for nicer labels (optional)

#Compute Spearman Correlation
spearman_global <- cor.test(
  comparison_df$signed_log10p_illumina,
  comparison_df$signed_log10p_pacbio,
  method = "spearman"
)
# Extract values
rho_value <- round(spearman_global$estimate, 2)
p_value <- signif(spearman_global$p.value, 2)

# Identify top 25 most divergent significant taxa for labeling
top_taxa <- comparison_df %>%
  filter(significant == "FDR < 0.05") %>%
  mutate(abs_diff = abs(signed_log10p_illumina - signed_log10p_pacbio)) %>%
  group_by(Taxon) %>%
  slice_max(order_by = abs_diff, n = 1) %>%   # pick most divergent per Taxon
  ungroup() %>%
  slice_max(order_by = abs_diff, n = 25)      # now select top 25 across all taxa

#Plot
p_signed_log10p <- ggplot(comparison_df, aes(x = signed_log10p_illumina, y = signed_log10p_pacbio)) +
  # Base points
  geom_point(data = subset(comparison_df, significant == "n.s."),
             color = "gray70", size = 1.2, alpha = 0.6) +
  geom_point(data = subset(comparison_df, significant == "FDR < 0.05"),
             color = "firebrick", size = 1.5, alpha = 0.9) +
  # Label top 10 divergent taxa only
  geom_text_repel(
    data = top_taxa,
    aes(label = Taxon),
    size = 3,
    box.padding = 0.4,
    point.padding = 0.2,
    segment.color = "gray60",
    segment.size = 0.3,
    force = 1.2,
    max.overlaps = Inf,
    fontface = "italic",    # or "plain" if names are not taxonomic
    color = "firebrick"
  ) +
  # Reference lines
  geom_vline(xintercept = 0, linetype = "dashed", color = "darkred") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  #geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "gray60") +
  #geom_smooth(method = "lm", color = "black", se = FALSE, linetype = "solid") +
  # Spearman annotation
  annotate(
    "text",
    x = min(comparison_df$signed_log10p_illumina, na.rm = TRUE),
    y = max(comparison_df$signed_log10p_pacbio, na.rm = TRUE),
    label = paste0("ρ = ", rho_value, "\np = ", format(p_value, scientific = TRUE)),
    hjust = 0, vjust = 1.2, size = 5
  ) +
  # Theme and labels
  theme_classic(base_size = 14) +
  labs(
    x = expression("Signed "*-log[10]*"(p) Illumina"),
    y = expression("Signed "*-log[10]*"(p) PacBio")
  )

p_signed_log10p

cor_results <- comparison_df %>%
  group_by(term) %>%
  summarise(
    spearman_r = cor(signed_log10p_illumina, signed_log10p_pacbio, method = "spearman"),
    p_value = cor.test(signed_log10p_illumina, signed_log10p_pacbio, method = "spearman")$p.value,
    .groups = "drop"
  )

print(cor_results)
```
##Fig 3b: P-value scatter: Compare locations
```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(broom)
library(purrr)
library(ggplot2)
library(ggrepel)

# Step 1: Define all pairwise combinations of locations
location_pairs <- combn(
  unique(overlapping_genus$Location) %>% setdiff("Control"),
  2,
  simplify = FALSE
)

# Step 2: Create function to run per-method pairwise comparisons
run_pairwise_comparisons <- function(df, method_col, method_label) {
  map_dfr(location_pairs, function(pair) {
    loc1 <- pair[1]
    loc2 <- pair[2]
    comp_label <- paste(loc1, "vs", loc2, sep = "_")

    # Subset and define reference group
    sub_df <- df %>%
      filter(Location %in% c(loc1, loc2)) %>%
      mutate(Location = factor(Location, levels = c(loc1, loc2)))  # loc1 = reference

    if (nrow(sub_df) == 0) return(tibble())  # skip empty groups

    sub_df %>%
      group_by(Taxon) %>%
      nest() %>%
      mutate(
        model = map(data, ~ tryCatch(
          lm(reformulate("Location", method_col), data = .x),
          error = function(e) NULL
        )),
        tidied = map(model, ~ if (!is.null(.x)) tidy(.x) else NULL)
      ) %>%
      unnest(tidied) %>%
      filter(term != "(Intercept)", !is.na(estimate), !is.na(p.value)) %>%
      mutate(
        signed_log10p = sign(estimate) * -log10(p.value),
        comparison_group = comp_label,
        method = method_label
      ) %>%
      select(Taxon, comparison_group, estimate, p.value, signed_log10p, method)
  })
}

# Step 3: Run the function for both pipelines
res_pacbio <- run_pairwise_comparisons(overlapping_genus, "logRelAbund_PacBio", "PacBio")
res_illumina <- run_pairwise_comparisons(overlapping_genus, "logRelAbund_Illumina", "Illumina")

# Step 4: Adjust p-values (FDR) within each comparison group
res_pacbio <- res_pacbio %>%
  group_by(comparison_group) %>%
  mutate(fdr_pacbio = p.adjust(p.value, method = "BH")) %>%
  ungroup()

res_illumina <- res_illumina %>%
  group_by(comparison_group) %>%
  mutate(fdr_illumina = p.adjust(p.value, method = "BH")) %>%
  ungroup()

# Step 5: Merge both results by Taxon + comparison group
comparison_df <- res_pacbio %>%
  rename(
    estimate_pacbio = estimate,
    log10p_pacbio = signed_log10p
  ) %>%
  inner_join(
    res_illumina %>%
      rename(
        estimate_illumina = estimate,
        log10p_illumina = signed_log10p
      ),
    by = c("Taxon", "comparison_group")
  ) %>%
  mutate(
    significant = ifelse(fdr_pacbio < 0.05 | fdr_illumina < 0.05, "FDR < 0.05", "n.s.")
  )

# Step 6: Define safe Spearman correlation function
safe_spearman <- function(x, y) {
  if (length(x) < 3 || length(unique(x)) == 1 || length(unique(y)) == 1) {
    return(c(rho = NA, p = NA))
  }
  test <- suppressWarnings(cor.test(x, y, method = "spearman"))
  return(c(rho = as.numeric(test$estimate), p = as.numeric(test$p.value)))
}

# Step 7: Compute correlation per group
cor_summary <- comparison_df %>%
  group_split(comparison_group) %>%
  map_dfr(function(df_group) {
    cg <- unique(df_group$comparison_group)
    
    if (!all(c("log10p_illumina", "log10p_pacbio") %in% colnames(df_group))) {
      message("Missing columns in: ", cg)
      return(tibble(comparison_group = cg, spearman_r = NA, spearman_p = NA))
    }

    if (sum(complete.cases(df_group$log10p_illumina, df_group$log10p_pacbio)) < 3) {
      message("Too few valid points in: ", cg)
      return(tibble(comparison_group = cg, spearman_r = NA, spearman_p = NA))
    }

    result <- safe_spearman(df_group$log10p_illumina, df_group$log10p_pacbio)
    tibble(comparison_group = cg,
           spearman_r = result["rho"],
           spearman_p = result["p"])
  }) %>%
  mutate(spearman_fdr = p.adjust(spearman_p, method = "BH"))

# Step 8: Annotate and plot
comparison_df_annotated <- comparison_df %>%
  left_join(cor_summary, by = "comparison_group")

# Step 9: Final plot
p_pvalue_group <- ggplot(comparison_df_annotated, aes(x = log10p_illumina, y = log10p_pacbio, color = significant)) +
  geom_point(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +
  geom_text_repel(
    data = filter(comparison_df_annotated, significant == "FDR < 0.05"),
    aes(label = Taxon), size = 3, max.overlaps = 100
  ) +
  facet_wrap(~comparison_group, scales = "free") +
  geom_text(
    data = cor_summary,
    aes(x = -Inf, y = Inf,
        label = paste0("ρ = ", round(spearman_r, 2),
                       "\nFDR = ", signif(spearman_fdr, 2))),
    inherit.aes = FALSE,
    hjust = -0.1, vjust = 1.2, size = 4
  ) +
  scale_color_manual(values = c("FDR < 0.05" = "firebrick", "n.s." = "skyblue")) +
  labs(
    title = "Signed –log₁₀(p-value) agreement across pipelines",
    x = expression("Signed "*-log[10]*"(p) (ITS2 – Illumina)"),
    y = expression("Signed "*-log[10]*"(p) (Full-length ITS – PacBio)"),
    color = "Significance"
  ) +
  theme_minimal(base_size = 14)

p_pvalue_group

#Summary plot from Spearman correlations
cor_summary <- cor_summary %>%
  mutate(sig_flag = ifelse(spearman_fdr < 0.05, "FDR < 0.05", "n.s."))

summary_plot <- cor_summary %>%
  ggplot(aes(x = reorder(comparison_group, spearman_r),
             y = spearman_r,
             fill = sig_flag)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(spearman_r, 2)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("FDR < 0.05" = "firebrick", "n.s." = "gray70")) +
  theme_bw(base_size = 14) +
  labs(
    title = "Agreement between sequencing pipelines across location comparisons",
    x = "Pairwise Comparison",
    y = expression("Spearman's " * rho),
    fill = "Significance"
  )
summary_plot 
```


##Function: clr transformer
```{r}
# CLR transformation helper
clr_transform <- function(ps, rank) {
  library(phyloseq)
  library(tidyverse)
  library(compositions)
  
  counts <- otu_table(ps)
  if (taxa_are_rows(ps)) counts <- t(counts)
  
  # Add pseudo-count to avoid log(0)
  counts <- counts + 1
  
  # CLR transform
  clr_mat <- clr(counts)
  df <- as.data.frame(clr_mat)
  df$SampleID <- rownames(df)
  
  # Convert to long format
  df_long <- pivot_longer(df, -SampleID, names_to = "FeatureID", values_to = "CLR")
  
  # Extract taxonomy
  tax <- as.data.frame(tax_table(ps))
  tax$FeatureID <- rownames(tax)
  
  # Join taxonomy at specified rank
  df_long <- left_join(df_long, tax[, c("FeatureID", rank)], by = "FeatureID")
  df_long <- df_long %>%
    rename(Taxon = all_of(rank)) %>%
    filter(!is.na(Taxon))  # Optionally drop unassigned

  return(df_long)
}
```

##Scatter plot
```{r}
# Store merged data per rank
merged_list <- list()

for (i in seq_along(rank_ITS)) {
  CLR_PacBio <- clr_transform(ps_PacBio_ITS_match_glom, rank = rank_ITS[i])
  CLR_Illumina <- clr_transform(ps_Illumina_ITS2_glom, rank = rank_ITS2[i])
  
  merged <- full_join(CLR_PacBio, CLR_Illumina, by = c("SampleID", "Taxon"), suffix = c("_PacBio", "_Illumina"))
  
  # NEW: Keep only rows where both CLR values are not NA
  merged <- merged %>% filter(!is.na(CLR_PacBio) & !is.na(CLR_Illumina))
  
  merged$Rank <- rank_ITS[i]
  merged_list[[rank_ITS[i]]] <- merged
}

# Combine all ranks into one data frame
merged_all <- bind_rows(merged_list)
```

#Fig 3a: Box plot: Spearman correlation of CLR-transformed abundance for each sample
```{r}
genus_data <- merged_all %>% filter(Rank == "Genus")

#Correlation data.frame
sample_correlations <- genus_data %>%
  group_by(SampleID) %>%
  summarize(
    Spearman_rho = cor(CLR_PacBio, CLR_Illumina, method = "spearman", use = "complete.obs"),
    n_taxa = n()
  ) %>%
  arrange(desc(Spearman_rho))

#Add grouping columns
metadata_clean <- metadata_ITS %>%
  select(Label, SampleType, Location)

sample_correlations <- left_join(sample_correlations, metadata_clean, by = c("SampleID" = "Label"))
summary(sample_correlations)

#Group by sample type
sample_correlations <- sample_correlations %>%
  mutate(SampleType = case_when(
    SampleType == "swab" ~ "Swab",
    SampleType == "air" ~ "Air",
    SampleType == "NegCon" ~ "NegCon",
    SampleType == "PosCon" ~ "PosCon"
  ))

sd(sample_correlations$Spearman_rho, na.rm = TRUE)
mean(sample_correlations$Spearman_rho, na.rm = TRUE)

shapiro.test(sample_correlations$Spearman_rho)

#Plot
box_Location <- ggplot(sample_correlations, aes(x = Location, y = Spearman_rho)) +
  geom_boxplot(fill = "lightgray", outlier.shape = NA) +
  geom_jitter(aes(color = Location), width = 0.2, alpha = 0.8, size = 2) +
  theme_classic(base_size = 14) +
  scale_color_brewer(palette = "Set2", drop = FALSE) +
  theme(panel.border = element_rect(color = "black", fill = NA), # Adds an outer frame
        axis.line = element_line(color = "black"),               # Keeps axis lines
        panel.grid = element_blank()) +                            # Ensures no inner grid lines
  labs(title = "Spearman correlation (Genus)",
       x = "Location", y = "Spearman Coefficient") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

box_Location

box_sampletype <- ggplot(sample_correlations, aes(x = SampleType, y = Spearman_rho)) +
  geom_boxplot(fill = "lightgray", outlier.shape = NA) +
  geom_jitter(aes(color = Location), width = 0.2, alpha = 0.8, size = 2) +
  theme_classic(base_size = 14) +
  scale_color_brewer(palette = "Set2", drop = FALSE) +
  theme(panel.border = element_rect(color = "black", fill = NA), # Adds an outer frame
        axis.line = element_line(color = "black"),               # Keeps axis lines
        panel.grid = element_blank()) +                            # Ensures no inner grid lines
  labs(title = "Spearman correlation (Genus)",
       x = "Sample Type", y = "Spearman Coefficient") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

box_sampletype
```
#Correlation of most abundant
```{r}
genus_abundance <- psmelt(ps_PacBio_ITS_match_glom) %>%
  group_by(Genus) %>%
  summarize(mean_abundance = mean(Abundance / sum(Abundance)))

top_genera <- genus_abundance %>%
  slice_head(n = 100) %>%
  pull(Genus)

genus_abundant <- genus_data %>%
  filter(Taxon %in% top_genera)

sample_correlations_abundant <- genus_abundant %>%
  group_by(SampleID) %>%
  summarize(
    Spearman_rho = cor(CLR_PacBio, CLR_Illumina,
                       method = "spearman",
                       use = "complete.obs"),
    n_taxa = n()
  ) %>%
  arrange(desc(Spearman_rho))

summary(sample_correlations_abundant$Spearman_rho)
```


#ANOVA spearman r
```{r}
library(car)
leveneTest(Spearman_rho ~ SampleType, data = sample_correlations)
leveneTest(Spearman_rho ~ Location, data = sample_correlations)

anova_type <- aov(Spearman_rho ~ SampleType, data = sample_correlations)
summary(anova_type)
TukeyHSD(anova_type)

anova_location <- aov(Spearman_rho ~ Location, data = sample_correlations)
summary(anova_location)
TukeyHSD(anova_location)

anova_all <- aov(Spearman_rho ~ Location + SampleType, data = sample_correlations)
summary(anova_all)
TukeyHSD(anova_all)
```
#Fig 3c: Health-related genera: 
## Prepare data
```{r}
bubble_data <- genus_data %>%
  left_join(metadata_ITS, by = c("SampleID" = "Label")) %>%
  pivot_longer(cols = c(CLR_PacBio, CLR_Illumina), 
               names_to = "Method", values_to = "CLR_Abundance")

bubble_data$Method[bubble_data$Method == "CLR_PacBio"] <- "Full-length ITS"
bubble_data$Method[bubble_data$Method == "CLR_Illumina"] <- "ITS2"

#Filter taxa of interest
taxa_of_interest <- c(
  # Pathogens
  "Candida", "Aspergillus", "Cryptococcus", "Fusarium", "Trichosporon", 
  "Rhizopus", "Mucor", "Scedosporium", "Histoplasma", "Coccidioides",
  
  # Food spoilage
  "Penicillium", "Cladosporium", "Alternaria", "Geotrichum", "Botrytis", 
  "Wallemia",
  
  # Indoor/environmental molds
  "Stachybotrys", "Chaetomium", "Acremonium", "Trichoderma", 
  "Epicoccum", "Ulocladium"
)

target_genera <- c("Bipolaris", "Blastomyces",
  "Botrytis", "Candida", "Chaetomium", "Clavispora", "Coccidioides",
  "Cryptococcus", "Debaryomyces", "Drechslera", "Eurotium", 
  "Histoplasma",  "Mucor", "Phoma", "Pichia", "Puccinia",
  "Rhizopus", "Rhodotorula", "Saccharomyces", "Scopulariopsis", "Stachybotrys","Stemphylium", "Epicoccum", "Aureobasidium","Trichophyton", "Ustilago", "Malassezia", "Wallemia", "Wickerhamomyces","Bjerkandera", "Trichoderma","Alternaria",  "Fusarium","Aspergillus", "Penicillium", "Cladosporium")

bubble_data <- bubble_data %>% filter(Taxon %in% target_genera)

#Plot by location
bubble_summary <- bubble_data %>%
  group_by(Taxon, Location, Method, SampleType) %>%
  summarize(
    CLR_Abundance = mean(CLR_Abundance, na.rm = TRUE),
    .groups = "drop"
  )

#Plot
# Reorder Taxon by target list
bubble_summary$Taxon <- factor(bubble_summary$Taxon, levels = target_genera)
```
##Heatmap
```{r}
library(pheatmap)
library(gridExtra)
library(RColorBrewer)

heatmap_summary <- bubble_summary %>%
  mutate(
    Location = factor(Location, levels = c("Home1","Home2","Dorm","Lab1", "Lab2","Control")),
    Method   = factor(Method,   levels = c("Full-length ITS","ITS2")),
    SampleType = factor(SampleType, levels = c("air","swab","PosCon","NegCon"))
  )

heatmap_summary <- heatmap_summary %>%
  mutate(LocMethod = paste(Location, Method, sep = "_"))

#wide matrix: rows = Taxon, cols = LocMethod
mat_loc <- heatmap_summary %>%
  select(Taxon, LocMethod, CLR_Abundance) %>%
  mutate(CLR_Abundance = suppressWarnings(as.numeric(CLR_Abundance))) %>%
  pivot_wider(
    names_from  = LocMethod,
    values_from = CLR_Abundance,
    values_fn   = list(CLR_Abundance = ~ mean(.x, na.rm = TRUE)),  # aggregate duplicates
    values_fill = 0
  ) %>%
  as.data.frame()

row.names(mat_loc) <- mat_loc$Taxon
mat_loc <- mat_loc[, -1]

# handle missing values
mat_loc[is.na(mat_loc)] <- 0

#Heatmap: by SampleType
heatmap_summary <- heatmap_summary %>%
  mutate(TypeMethod = paste(SampleType, Method, sep = "_"))

#wide matrix: rows = Taxon, cols = LocMethod
mat_type <- heatmap_summary %>%
  select(Taxon, TypeMethod, CLR_Abundance) %>%
  mutate(CLR_Abundance = suppressWarnings(as.numeric(CLR_Abundance))) %>%
  pivot_wider(
    names_from  = TypeMethod,
    values_from = CLR_Abundance,
    values_fn   = list(CLR_Abundance = ~ mean(.x, na.rm = TRUE)),  # aggregate duplicates
    values_fill = 0
  ) %>%
  as.data.frame()

row.names(mat_type) <- mat_type$Taxon
mat_type <- mat_type[, -1]

# handle missing values
mat_type[is.na(mat_type)] <- 0

#Set color scale
#Global range
rng  <- range(c(mat_loc, mat_type), na.rm = TRUE)   # c(-0.282, 11.252)
gmin <- rng[1]
gmax <- rng[2]

# Shared breaks & colors (sequential)
n_breaks <- 100
breaks   <- seq(gmin, gmax, length.out = n_breaks + 1)  # 101 edges for 100 color bins
clr_pal  <- colorRampPalette(c("#F7F7F7", "#B2182B"))(n_breaks)

# heatmap: by location
## column annotations: Location & Method
ann_col_loc <- data.frame(
  Location = str_split_fixed(colnames(mat_loc), "_", 2)[ , 1],
  Method   = str_split_fixed(colnames(mat_loc), "_", 2)[ , 2]
)
rownames(ann_col_loc) <- colnames(mat_loc)

loc_cols <- c("Control" = brewer.pal(8, "Set2")[1],
              "Dorm" = brewer.pal(8, "Set2")[2],
              "Home1" = brewer.pal(8, "Set2")[3],
              "Home2" = brewer.pal(8, "Set2")[4],
              "Lab1" = brewer.pal(8, "Set2")[5],
              "Lab2" = brewer.pal(8, "Set2")[6])

method_cols <- c("Full-length ITS" = brewer.pal(8, "Set2")[7],
                 "ITS2"  = brewer.pal(8, "Set2")[8])

ann_colors_loc <- list(Location = loc_cols, Method = method_cols)

##Plot
heatmap_byLocation <- pheatmap(mat_loc, cellwidth = 10, cellheight = 10, color = clr_pal, breaks = breaks, cluster_rows = FALSE, cluster_cols = FALSE, border_color = NA, display_numbers = FALSE, fontsize_row = 9, fontsize_col = 8, main = "CLR Abundance of Health‑related Genera", annotation_col = ann_col_loc, annotation_colors = ann_colors_loc, legend = TRUE)

heatmap_byLocation

# heatmap: by location
# column annotations: Location & Method
ann_col_type <- data.frame(
  SampleType = str_split_fixed(colnames(mat_type), "_", 2)[ , 1],
  Method   = str_split_fixed(colnames(mat_type), "_", 2)[ , 2]
)
rownames(ann_col_type) <- colnames(mat_type)

type_cols <- c("NegCon" = brewer.pal(8, "Set1")[1],
              "PosCon" = brewer.pal(8, "Set1")[2],
              "swab" = brewer.pal(8, "Set1")[3],
              "air" = brewer.pal(8, "Set1")[4])

method_cols <- c("Full-length ITS" = brewer.pal(8, "Set2")[7],
                 "ITS2"  = brewer.pal(8, "Set2")[8])

ann_colors_type <- list(SampleType = type_cols, Method = method_cols)

#Order samples
desired <- c("NegCon_Full-length ITS", "NegCon_ITS2", "PosCon_Full-length ITS", "PosCon_ITS2", "swab_Full-length ITS", "swab_ITS2", "air_Full-length ITS","air_ITS2")

#Reorder data
mat_type <- mat_type[, desired]
ann_col_type <- ann_col_type[desired, ]

##Plot
heatmap_bySampleType <- pheatmap(mat_type, cellwidth = 10, cellheight = 10, color = clr_pal, breaks = breaks, cluster_rows = FALSE, cluster_cols = FALSE, border_color = NA, display_numbers = FALSE, fontsize_row = 9, fontsize_col = 8, main = "CLR Abundance of Health-related Genera", annotation_col = ann_col_type, annotation_colors = ann_colors_type, legend = TRUE)

heatmap_bySampleType
```

## Statistics: wilcox test
```{r}
library(dplyr)
#Normality test
clr_wide <- bubble_data %>%
  select(SampleID, Taxon, Location, SampleType, Method, CLR_Abundance) %>%
  pivot_wider(names_from = Method, values_from = CLR_Abundance)

results <- clr_wide %>%
  group_by(Taxon) %>%
  summarize(
    n = n(),
    shapiro_p_ITS = if (n() >= 3) shapiro.test(`Full-length ITS`)$p.value else NA,
    shapiro_p_ITS2 = if (n() >= 3) shapiro.test(ITS2)$p.value else NA,
    test = case_when(
      !is.na(shapiro_p_ITS) & !is.na(shapiro_p_ITS2) &
      shapiro_p_ITS > 0.05 & shapiro_p_ITS2 > 0.05 ~ "t-test",
      TRUE ~ "wilcox"
    ),
    p_value = case_when(
      test == "t-test" ~ t.test(`Full-length ITS`, ITS2, paired = TRUE)$p.value,
      test == "wilcox" ~ wilcox.test(`Full-length ITS`, ITS2, paired = TRUE)$p.value,
      TRUE ~ NA_real_
    )
  ) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))  # FDR correction

#Wilcox-test to compare full-length ITS vs ITS2
wilcox_results <- clr_wide %>%
  group_by(Taxon) %>%
  summarize(
    n = n(),
    p_value = tryCatch(
      wilcox.test(`Full-length ITS`, ITS2, paired = TRUE)$p.value,
      error = function(e) NA
    )
  ) %>%
  ungroup() %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),  # adjust for multiple testing
    significance = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    )
  )

#compare within sample type or location
library(dplyr)
library(purrr)
library(tidyr)

# Step 1: Group and nest data by location
wilcox_results_location <- clr_wide %>%
  group_by(Taxon, Location) %>%
  filter(!is.na(`Full-length ITS`) & !is.na(ITS2)) %>%  # remove NAs
  nest() %>%
  mutate(
    test_result = map(data, ~ tryCatch(
      wilcox.test(.x$`Full-length ITS`, .x$ITS2, paired = TRUE),
      error = function(e) NULL
    )),
    p_value = map_dbl(test_result, ~ ifelse(is.null(.x), NA, .x$p.value))
  ) %>%
  ungroup() %>%
  select(Taxon, Location, p_value) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),  # adjust for multiple testing
    significance = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    )
  )

# Step 2: Group and nest data by sample type
wilcox_results_sampletype <- clr_wide %>%
  group_by(Taxon, SampleType) %>%
  filter(!is.na(`Full-length ITS`) & !is.na(ITS2)) %>%  # remove NAs
  nest() %>%
  mutate(
    test_result = map(data, ~ tryCatch(
      wilcox.test(.x$`Full-length ITS`, .x$ITS2, paired = TRUE),
      error = function(e) NULL
    )),
    p_value = map_dbl(test_result, ~ ifelse(is.null(.x), NA, .x$p.value))
  ) %>%
  ungroup() %>%
  select(Taxon, SampleType, p_value) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),  # adjust for multiple testing
    significance = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ ""
    )
  )

wilcox_results_sampletype

```

##Scatter plot: correlation
```{r}
# Set desired order here
desired_location_order <- c("Control","Dorm", "Home1", "Home2", "Lab1", "Lab2")
clr_wide$Location <- factor(clr_wide$Location, levels = desired_location_order)

# Step 2: Scatter plot of CLR abundances between platforms
scatter_healthtaxa <- ggplot(clr_wide, aes(x = `Full-length ITS`, y = ITS2, color = Location)) +
  geom_point(alpha = 0.8, size = 2) +
   geom_smooth(method = "lm", color = "gray20", se = FALSE) +
   geom_abline(slope = 1, intercept = 0, linetype = "dotdash", color = "gray70") +
   scale_color_brewer(palette = "Set2", drop = FALSE) +
   facet_wrap(~SampleType) +
   #geom_text_repel(data = filter(clr_wide, Taxon %in% sig_taxa), aes(label = Taxon), size = 3, max.overlaps = Inf, box.padding = 0.3, point.padding = 0.3) +
  stat_cor(
    method = "spearman",
    aes(label = paste(..rr.label.., sep = "~`,`~")),
    size = 5, label.x = 0.5
  ) +
  labs(
    x = "CLR Abundance (Full-length ITS)",
    y = "CLR Abundance (ITS2)",
    title = "Correlation of Genus-Level CLR Abundance Across Platforms",
    subtitle = "Each point is a genus in a sample"
  ) +
  theme_bw(base_size = 14)

scatter_healthtaxa
```


#Fig 1b: Top 20 genera plot (combined mean CLR)
```{r}
# Step 1: summarize CLR distribution per genus × platform
genus_summary <- genus_data %>%
  pivot_longer(cols = c(CLR_Illumina, CLR_PacBio),
               names_to = "Method",
               values_to = "CLR") %>%
  mutate(Method = ifelse(Method == "CLR_Illumina", "ITS2", "Full-length ITS")) %>%
  group_by(Taxon, Method) %>%
  summarize(
    mean_CLR = mean(CLR, na.rm = TRUE),
    se_CLR   = sd(CLR, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Step 2: choose top 20 genera by mean across both pipelines
top20_taxa <- genus_summary %>%
  group_by(Taxon) %>%
  summarize(mean_across = mean(mean_CLR)) %>%
  arrange(desc(mean_across)) %>%
  slice_head(n = 20) %>%
  pull(Taxon)

top20_plotdata <- genus_summary %>% filter(Taxon %in% top20_taxa)

# Step 3: plot with error bars
Barplot_combinedtop20 <- ggplot(top20_plotdata,
       aes(x = reorder(Taxon, -mean_CLR), y = mean_CLR, fill = Method)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7, color = "black", linewidth = 0.5) +
  geom_errorbar(aes(ymin = mean_CLR - se_CLR, ymax = mean_CLR + se_CLR),
                position = position_dodge(width = 0.7),
                width = 0.25, linewidth = 0.5) +
  theme_classic(base_size = 14) +
  labs(
    title = "Top 20 Genera (Mean CLR Abundance ± SE)",
    x = "Genus",
    y = "Mean CLR (centered log-ratio)"
  ) +
  scale_fill_manual(values = c("ITS2" = "orange", "Full-length ITS" = "deeppink")) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA),
        axis.line = element_line(color = "black"))

Barplot_combinedtop20
```

#Fig S5: Top 20 genera plots (separate mean CLR)
```{r}
# Pick top 20 genera by ITS2's mean CLR
top20_ITS2 <- genus_summary %>%
  filter(Method == "ITS2") %>%
  arrange(desc(mean_CLR)) %>%
  slice_head(n = 20) %>%
  pull(Taxon)

plot_dat_ITS2 <- genus_summary %>%
  filter(Taxon %in% top20_ITS2) %>%
  mutate(Taxon = factor(Taxon, levels = top20_ITS2))

top20_by_ITS2 <- ggplot(plot_dat_ITS2,
       aes(x = Taxon, y = mean_CLR, fill = Method)) +
  geom_col(position = position_dodge(width = 0.7),
           width = 0.7, color = "black", linewidth = 0.3) +
  geom_errorbar(aes(ymin = mean_CLR - se_CLR, ymax = mean_CLR + se_CLR),
                position = position_dodge(width = 0.7),
                width = 0.25, linewidth = 0.3) +
  scale_fill_manual(values = c("ITS2" = "orange", "Full-length ITS" = "deeppink")) +
  labs(title = "Top 20 Genera Ranked by ITS2 (Mean CLR ± SE)",
       x = "Genus", y = "Mean CLR (centered log-ratio)", fill = "Method") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA),
        axis.line = element_line(color = "black"))

top20_by_ITS2

# Pick top 20 genera by Full-length ITS mean CLR
top20_FullITS <- genus_summary %>%
  filter(Method == "Full-length ITS") %>%
  arrange(desc(mean_CLR)) %>%
  slice_head(n = 20) %>%
  pull(Taxon)

plot_dat_FullITS <- genus_summary %>%
  filter(Taxon %in% top20_FullITS) %>%
  mutate(Taxon = factor(Taxon, levels = top20_FullITS))

top20_by_FullITS <- ggplot(plot_dat_FullITS,
       aes(x = Taxon, y = mean_CLR, fill = Method)) +
  geom_col(position = position_dodge(width = 0.7),
           width = 0.7, color = "black", linewidth = 0.3) +
  geom_errorbar(aes(ymin = mean_CLR - se_CLR, ymax = mean_CLR + se_CLR),
                position = position_dodge(width = 0.7),
                width = 0.25, linewidth = 0.3) +
  scale_fill_manual(values = c("ITS2" = "orange", "Full-length ITS" = "deeppink")) +
  labs(title = "Top 20 Genera Ranked by Full-length ITS (Mean CLR ± SE)",
       x = "Genus", y = "Mean CLR (centered log-ratio)", fill = "Method") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA),
        axis.line = element_line(color = "black"))

top20_by_FullITS
```


#PacBio vs Illumina comparison
#Fig 2a: Alpha-diversity group by sample type
```{r}
# 1. Compute alpha diversity for each pipeline
ITS2_alpha <- estimate_richness(ps_Illumina_ITS2_filt,
                                measures = c("Shannon", "Chao1", "Observed")) %>%
  mutate(Label = rownames(.),
         Pipeline = "ITS2")


ITS_alpha <- estimate_richness(ps_PacBio_ITS_filt,
                                   measures = c("Shannon", "Chao1", "Observed")) %>%
  mutate(Label = rownames(.),
         Pipeline = "Full-length ITS")

# 2. Inner join on Label to ensure true pairs
alpha_pairs <- inner_join(
  ITS2_alpha %>% select(Label, Shannon_ITS2 = Shannon,
                        Chao1_ITS2 = Chao1, Obs_ITS2 = Observed),
  ITS_alpha %>% select(Label, Shannon_ITS = Shannon,
                           Chao1_ITS = Chao1, Obs_ITS = Observed),
  by = "Label"
)

metadata <- data.frame(ps_Illumina_ITS2_filt@sam_data)
alpha_pairs <- left_join(alpha_pairs, metadata, by = "Label")

# Reshape the dataframe
alpha_long <- alpha_pairs %>%
  transmute(
    Label, SampleType, Location,
    Shannon_ITS2 = Shannon_ITS2, Shannon_ITS = Shannon_ITS,
    Chao1_ITS2   = Chao1_ITS2,   Chao1_ITS   = Chao1_ITS,
    Observed_ITS2 = Obs_ITS2,    Observed_ITS = Obs_ITS
  ) %>%
  pivot_longer(
    cols = -c(Label, SampleType, Location),
    names_to  = c("Metric","Pipeline"),
    names_sep = "_",
    values_to = "Value"
  ) %>%
  mutate(Pipeline = ifelse(Pipeline == "ITS2", "ITS2", "Full-length ITS"))

# 3. Paired Wilcoxon tests
paired_alpha_tests <- alpha_long %>%
  pivot_wider(id_cols = c(Label, SampleType, Metric),
              names_from = Pipeline,
              values_from = Value) %>%
  filter(!is.na(`Full-length ITS`), !is.na(ITS2)) %>%
  group_by(SampleType, Metric) %>%
  summarise(
    n           = n(),
    median_diff = median(`Full-length ITS` - ITS2, na.rm = TRUE),
    p_value     = wilcox.test(`Full-length ITS`, ITS2, paired = TRUE,
                              exact = FALSE)$p.value,
    W           = wilcox.test(`Full-length ITS`, ITS2, paired = TRUE,
                              exact = FALSE)$statistic,
    .groups = "drop"
  ) %>%
  group_by(Metric) %>%                # FDR across SampleTypes within each metric
  mutate(q_value = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  mutate(sig = case_when(
    is.na(q_value)      ~ "",
    q_value < 0.001     ~ "***",
    q_value < 0.01      ~ "**",
    q_value < 0.05      ~ "*",
    TRUE                ~ "ns"
  ))

#4. Plot: paired box with lines
desired_location_order <- c("Control","Dorm", "Home1", "Home2", "Lab1", "Lab2")
alpha_long$Location <- factor(alpha_long$Location, levels = desired_location_order)

plot_alpha_paired_facet_by_loc <- function(metric_name, ylab) {

  df_plot <- alpha_long %>% 
    filter(Metric == metric_name) %>%
    # ensure Location exists (pulled from your metadata join)
    mutate(Location = factor(Location))      

  # y-position per facet (just above max in that facet)
  y_max <- df_plot %>%
    group_by(SampleType) %>%
    summarise(y = max(Value, na.rm = TRUE), .groups = "drop")

  ann <- paired_alpha_tests %>%
    filter(Metric == metric_name) %>%
    select(SampleType, sig, q_value) %>%
    left_join(y_max, by = "SampleType") %>%
    mutate(y = y + 0.05 * abs(y),   # headroom for the annotation
           x = 1.5)

  ggplot(df_plot, aes(x = Pipeline, y = Value, group = Label)) +
     # side-by-side boxplots per Pipeline (keep neutral fill)
    geom_boxplot(aes(group = Pipeline),
                 width = 0.6, fill = "grey88", color = "black",
                 outlier.shape = NA) +
    # neutral connecting lines emphasize pairing (not method or location)
    geom_line(alpha = 0.35, color = "grey55") +
    # points colored by Location (Dark2)
    geom_point(aes(color = Location), size = 2, alpha = 0.9,
               position = position_jitter(width = 0.04, height = 0)) +
    # significance label per facet (paired Wilcoxon within SampleType)
    geom_text(data = ann, aes(x = x, y = y, label = sig),
              inherit.aes = FALSE, size = 5) +
    scale_color_brewer(palette = "Set2", drop = FALSE) +  # <-- Dark2
    facet_wrap(~ SampleType, scales = "fixed") +
    labs(x = "", y = ylab,
         color = "Location",
         caption = "Paired Wilcoxon within SampleType; labels: ***<0.001, **<0.01, *<0.05, ns≥0.05") +
    theme_classic(base_size = 14) +
    theme(
      panel.border = element_rect(color = "black", fill = NA),
      axis.text.x  = element_text(angle = 45, hjust = 1)
    )
}

# Build your three panels
p_shannon  <- plot_alpha_paired_facet_by_loc("Shannon",  "Shannon diversity")
p_chao1    <- plot_alpha_paired_facet_by_loc("Chao1",    "Chao1 richness")
p_observed <- plot_alpha_paired_facet_by_loc("Observed", "Observed ASVs")

p_shannon; p_chao1; p_observed

```
## Alpha-diversity by Location
```{r}
# 3. Paired Wilcoxon tests
paired_alpha_tests <- alpha_long %>%
  pivot_wider(id_cols = c(Label, Location, Metric),
              names_from = Pipeline,
              values_from = Value) %>%
  filter(!is.na(`Full-length ITS`), !is.na(ITS2)) %>%
  group_by(Location, Metric) %>%
  summarise(
    n           = n(),
    median_diff = median(`Full-length ITS` - ITS2, na.rm = TRUE),
    p_value     = wilcox.test(`Full-length ITS`, ITS2, paired = TRUE,
                              exact = FALSE)$p.value,
    W           = wilcox.test(`Full-length ITS`, ITS2, paired = TRUE,
                              exact = FALSE)$statistic,
    .groups = "drop"
  ) %>%
  group_by(Metric) %>%                # FDR across SampleTypes within each metric
  mutate(q_value = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  mutate(sig = case_when(
    is.na(q_value)      ~ "",
    q_value < 0.001     ~ "***",
    q_value < 0.01      ~ "**",
    q_value < 0.05      ~ "*",
    TRUE                ~ "ns"
  ))

paired_alpha_tests

#4. Plot: paired box with lines
desired_location_order <- c("Control","Dorm", "Home1", "Home2", "Lab1", "Lab2")
alpha_long$Location <- factor(alpha_long$Location, levels = desired_location_order)

plot_alpha_paired_facet_by_loc <- function(metric_name, ylab) {

  df_plot <- alpha_long %>% 
    filter(Metric == metric_name) %>%
    # ensure Location exists (pulled from your metadata join)
    mutate(SampleType = factor(SampleType))      

  # y-position per facet (just above max in that facet)
  y_max <- df_plot %>%
    group_by(Location) %>%
    summarise(y = max(Value, na.rm = TRUE), .groups = "drop")

  ann <- paired_alpha_tests %>%
    filter(Metric == metric_name) %>%
    select(Location, sig, q_value) %>%
    left_join(y_max, by = "Location") %>%
    mutate(y = y + 0.05 * abs(y),   # headroom for the annotation
           x = 1.5)

  ggplot(df_plot, aes(x = Pipeline, y = Value, group = Label)) +
     # side-by-side boxplots per Pipeline (keep neutral fill)
    geom_boxplot(aes(group = Pipeline),
                 width = 0.6, fill = "grey88", color = "black",
                 outlier.shape = NA) +
    # neutral connecting lines emphasize pairing (not method or location)
    geom_line(alpha = 0.35, color = "grey55") +
    # points colored by Location (Dark2)
    geom_point(aes(color = Location), size = 2, alpha = 0.9,
               position = position_jitter(width = 0.04, height = 0)) +
    # significance label per facet (paired Wilcoxon within SampleType)
    geom_text(data = ann, aes(x = x, y = y, label = sig),
              inherit.aes = FALSE, size = 5) +
    scale_color_brewer(palette = "Set2", drop = FALSE) +  # <-- Dark2
    facet_wrap(~ Location, scales = "fixed") +
    labs(x = "", y = ylab,
         color = "Location",
         caption = "Paired Wilcoxon within SampleType; labels: ***<0.001, **<0.01, *<0.05, ns≥0.05") +
    theme_classic(base_size = 14) +
    theme(
      panel.border = element_rect(color = "black", fill = NA),
      axis.text.x  = element_text(angle = 45, hjust = 1)
    )
}

# Build your three panels
p_shannon  <- plot_alpha_paired_facet_by_loc("Shannon",  "Shannon diversity")
p_chao1    <- plot_alpha_paired_facet_by_loc("Chao1",    "Chao1 richness")
p_observed <- plot_alpha_paired_facet_by_loc("Observed", "Observed ASVs")

p_shannon; p_chao1; p_observed
```

##Fig S6: Correlation: Shannon
```{r}
# Pearson correlation
cor_shannon <- cor(alpha_pairs$Shannon_ITS2,
                   alpha_pairs$Shannon_ITS,
                   method = "spearman",
                   use = "complete.obs")

# R-squared
r2_shannon <- cor_shannon^2

ggplot(alpha_pairs,
       aes(x = Shannon_ITS2, y = Shannon_ITS)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey40") +
  annotate("text",
           x = min(alpha_pairs$Shannon_ITS2, na.rm = TRUE),
           y = max(alpha_pairs$Shannon_ITS, na.rm = TRUE),
           hjust = 0, vjust = 1,
           label = paste0("R² = ", round(r2_shannon, 2),
                          "\nr = ", round(cor_shannon, 2))) +
  labs(x = "Shannon diversity (ITS2)",
       y = "Shannon diversity (Full-length ITS)",
       title = "Paired Shannon diversity comparison") +
  theme_classic(base_size = 14)
```
##Fig S6: Correlation: Chao1
```{r}
# Pearson correlation
cor_chao1 <- cor(alpha_pairs$Chao1_ITS2,
                 alpha_pairs$Chao1_ITS,
                 method = "spearman",
                 use = "complete.obs")

# R-squared
r2_chao1 <- cor_chao1^2

ggplot(alpha_pairs,
       aes(x = Chao1_ITS2, y = Chao1_ITS)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey40") +
  annotate("text",
           x = min(alpha_pairs$Chao1_ITS2, na.rm = TRUE),
           y = max(alpha_pairs$Chao1_ITS, na.rm = TRUE),
           hjust = 0, vjust = 1,
           label = paste0("R² = ", round(r2_chao1, 2),
                          "\nr = ", round(cor_chao1, 2))) +
  labs(x = "Chao1 richness (ITS2)",
       y = "Chao1 richness (Full-length ITS)",
       title = "Paired richness comparison") +
  theme_classic(base_size = 14)
```


#Use PacBio ITS as reference database to annotate Illumina BLAST Hits with PacBio Taxonomy
##Optional: Create PacBio ITS reference database
```{r}
library(Biostrings)

# Extract refseq
refseqs <- refseq(ps_PacBio_ITS_filt)

# Write to FASTA
writeXStringSet(refseqs, filepath = "PacBio/pacbio_filt_refseq.fasta")

# Extract taxonomy table
tax <- tax_table(ps_PacBio_ITS_filt)

# Convert to data frame and add sequence IDs
tax_df_ITS <- as.data.frame(tax)
tax_df_ITS$SequenceID <- names(refseqs)

# Reorder columns: ID first
tax_df_ITS <- tax_df_ITS[, c("SequenceID", setdiff(colnames(tax_df_ITS), "SequenceID"))]

# Write to TSV
write.table(tax_df_ITS, file = "PacBio/pacbio_filt_taxonomy.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

##Get Illumina ITS2 sequence
```{r}
library(Biostrings)

# Extract refseq
refseqs <- refseq(ps_Illumina_ITS2_filt)

# Write to FASTA
writeXStringSet(refseqs, filepath = "Illumina/Illumina_filt_refseq.fasta")

# Extract taxonomy table
tax <- tax_table(ps_Illumina_ITS2_filt)

# Convert to data frame and add sequence IDs
tax_df_ITS2 <- as.data.frame(tax)
tax_df_ITS2$SequenceID <- names(refseqs)

# Reorder columns: ID first
tax_df_ITS2 <- tax_df_ITS2[, c("SequenceID", setdiff(colnames(tax_df_ITS2), "SequenceID"))]

# Write to TSV
write.table(tax_df_ITS2, file = "Illumina/Illumina_filt_taxonomy.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

#Fig 4b: Blast result: mapped ASVs
```{r}
# Load libraries
library(dplyr)
library(readr)

# Step 1: Load BLAST hits
blast_hits <- read_tsv("Illumina/Illumina_blast_hits_allpacbio.tsv", col_names = c(
  "qseqid", "sseqid", "pident", "length", "mismatch", "gapopen",
  "qstart", "qend", "sstart", "send", "evalue", "bitscore"
))

seq_asv_map <- tibble(
  SequenceID = names(refseq(ps_Illumina_ITS2_filt)),
  Sequence = as.character(refseq(ps_Illumina_ITS2_filt))
)

blast_hits_filt <- blast_hits %>%
  filter(
    qseqid %in% seq_asv_map$Sequence,
    sseqid %in% tax_df_ITS$SequenceID
  )

blast_hits_filt$qseqid_sequence <-blast_hits_filt$qseqid
blast_hits_filt$qseqid <- seq_asv_map$SequenceID[
  match(blast_hits_filt$qseqid_sequence, seq_asv_map$Sequence)
]

#Mapped hits
matched_hits <- blast_hits_filt %>%
  filter(pident >= 99, bitscore > 0)

matched_summary <- matched_hits %>%
  count(qseqid, sseqid)

multi_hits <- matched_summary %>%
  group_by(qseqid) %>%
  summarise(n_matches = n()) %>%
  filter(n_matches > 1)

nrow(multi_hits)  # Number of Illumina ASVs mapping to >1 PacBio ASV

total_hits <- matched_summary %>%
  group_by(qseqid) %>%
  summarise(n_matches = n()) %>%
  filter(n_matches >= 1)

nrow(total_hits)  # Number of Illumina ASVs mapping to >=1 PacBio ASV

total_hits <- matched_summary %>%
  group_by(sseqid) %>%
  summarise(n_matches = n()) %>%
  filter(n_matches >= 1)

nrow(total_hits) # Number of PacBio ASVs mapped by at least 1 ITS2 ASV

total_hits <- matched_summary %>%
  group_by(sseqid) %>%
  summarise(n_matches = n()) %>%
  filter(n_matches > 1)

nrow(total_hits) # Number of PacBio ASVs mapped by at least 1 ITS2 ASV

# Generate bar plot of mapping distribution: ITS2
top_hits <- matched_hits %>%
  filter(pident == 100, bitscore > 0)

mapping_counts <- top_hits %>%
  group_by(qseqid) %>%
  summarize(
    n_mappings = n_distinct(sseqid),
    .groups = "drop"
  )

mapping_distribution <- mapping_counts %>%
  count(n_mappings)

MappingDistribution_ITS2 <- ggplot(mapping_distribution, aes(x = factor(n_mappings), y = n)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  labs(
    title = "Distribution of ITS2 to Full-length ITS Mapping",
    x = "Number of Full-length ITS Matches per ITS2 ASV",
    y = "Number of ITS2 ASVs"
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank()
  )

MappingDistribution_ITS2

# Generate bar plot of mapping distribution: full-length ITS
mapping_counts <- top_hits %>%
  group_by(sseqid) %>%
  summarize(
    n_mappings = n_distinct(qseqid),
    .groups = "drop"
  )

mapping_distribution <- mapping_counts %>%
  count(n_mappings)

MappingDistribution_ITS <- ggplot(mapping_distribution, aes(x = factor(n_mappings), y = n)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  labs(
    title = "Distribution of Full-length ITS Mapping",
    x = "Number of ITS2 ASVs Matches per full-length ITS sequence",
    y = "Number of Full-length ITS ASVs"
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank()
  )

MappingDistribution_ITS
```


##Collapse 100% matched ITS ASVs
```{r}
#Extract all 100% hits
hits_100 <- matched_hits %>%
  filter(pident >= 99.999)

tax_ranks <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

#Add taxonomy from tax_df_ITS
hits_100_tax <- hits_100 %>%
  left_join(
    tax_df_ITS %>% select(SequenceID, all_of(tax_ranks)),
    by = c("sseqid" = "SequenceID")
  )

#Function to collapse
collapse_lca <- function(df, ranks) {
  out <- setNames(as.list(rep(NA_character_, length(ranks))), ranks)

  for (r in ranks) {
    vals <- unique(na.omit(df[[r]]))

    if (length(vals) == 1) {
      out[[r]] <- vals
    } else {
      break
    }
  }

  as_tibble(out)
}

#collapse taxonomy for each ITS2 ASV
collapsed_taxonomy <- hits_100_tax %>%
  group_by(qseqid) %>%
  group_modify(~ collapse_lca(.x, tax_ranks)) %>%
  ungroup()

#add useful metadata
hit_counts <- hits_100_tax %>%
  group_by(qseqid) %>%
  summarize(
    n_100_hits = n_distinct(sseqid),
    .groups = "drop"
  )

collapsed_taxonomy <- collapsed_taxonomy %>%
  left_join(hit_counts, by = "qseqid") %>%
  mutate(
    Match_Type = ifelse(n_100_hits == 1, "Unique 100% match", "Multiple 100% matches")
  )

# keep the original matched IDs
matched_ids <- hits_100_tax %>%
  group_by(qseqid) %>%
  summarize(
    matched_sseqids = paste(sort(unique(sseqid)), collapse = ";"),
    .groups = "drop"
  )

collapsed_taxonomy <- collapsed_taxonomy %>%
  left_join(matched_ids, by = "qseqid")

```

##Subset to mapped and unmapped phyloseq
```{r}
ps_Illumina_ITS2_mapped <- subset_taxa(ps_Illumina_ITS2_filt, taxa_names(ps_Illumina_ITS2_filt) %in% hits_100$qseqid)
ps_Illumina_ITS2_unmapped <- subset_taxa(ps_Illumina_ITS2_filt, !(taxa_names(ps_Illumina_ITS2_filt) %in% hits_100$qseqid))

ps_PacBio_Fungi_match_mapped <- subset_taxa(ps_PacBio_ITS_filt, taxa_names(ps_PacBio_ITS_filt) %in% hits_100$sseqid)
ps_PacBio_Fungi_match_unmapped <- subset_taxa(ps_PacBio_ITS_filt, !(taxa_names(ps_PacBio_ITS_filt) %in% hits_100$sseqid))

ps_Illumina_ITS2_mapped
ps_PacBio_Fungi_match_mapped

library(Biostrings)
Illumina_ITS2_mapped_seq <- refseq(ps_Illumina_ITS2_mapped)
writeXStringSet(Illumina_ITS2_mapped_seq, filepath = "Illumina/matched_ITS2_filt_refseqs.fasta", format = "fasta")
PacBio_ITS_mapped_seq <- refseq(ps_PacBio_Fungi_match_mapped)
writeXStringSet(PacBio_ITS_mapped_seq, filepath = "PacBio/matched_ITS_filt_refseqs.fasta", format = "fasta")

Illumina_ITS2_mapped_tax <- data.frame(ps_Illumina_ITS2_mapped@tax_table)
PacBio_ITS_mapped_tax <- data.frame(ps_PacBio_Fungi_match_mapped@tax_table)
```

##Fig 5a: Tree plot of mapped ASVs
```{r}
library(ggtree)
library(treeio)
library(gridExtra)
library(patchwork)  # for side-by-side plots

# Load trees
tree_illumina <- read.tree("Illumina/Illumina_ITS2_tree.nwk")
tree_pacbio <- read.tree("PacBio/PacBio_ITS_tree.nwk")

# Convert rownames to a column for joining
illumina_tax <- Illumina_ITS2_mapped_tax %>%
 rownames_to_column(var = "label")  # 'label' must match tip labels in the tree

pacbio_tax <- PacBio_ITS_mapped_tax %>%
  rownames_to_column(var = "label")

#Check label matching
setdiff(tree_illumina$tip.label, illumina_tax$label)
all(tree_illumina$tip.label %in% illumina_tax$label)

setdiff(tree_pacbio$tip.label, pacbio_tax$label)
all(tree_pacbio$tip.label %in% pacbio_tax$label)

#Phylum color settings
all_phyla <- union(unique(illumina_tax$phylum), unique(pacbio_tax$Phylum))
phylum_colors <- setNames(
  brewer.pal(n = max(3, length(all_phyla)), name = "Dark2")[1:length(all_phyla)],
  all_phyla
)

all_class <- union(unique(illumina_tax$class), unique(pacbio_tax$Class))
library(colorspace)
class_colors <- setNames(
  qualitative_hcl(length(all_class), palette = "Dark 3"),  # or "Set 3", "Harmonic"
  all_class
)

#Taxa to label: using health-related genera list
## Extract ASVs from key genera
illumina_labels_to_label <- illumina_tax[illumina_tax$genus %in% target_genera, ]$label
pacbio_labels_to_label <- pacbio_tax[pacbio_tax$Genus %in% target_genera, ]$label

# Plot with same scale
p1 <- ggtree(tree_illumina, aes(color = phylum)) %<+% illumina_tax +
  geom_tree(size = 0.2) +
  geom_tiplab(data = function(x) subset(x, label %in% illumina_labels_to_label), aes(label = genus), size = 4, geom = "text", align = TRUE, linesize = 0.1, show.legend = FALSE) +
  scale_color_manual(values = phylum_colors, na.value = "gray40", name = "Phylum") +
  theme_tree2() +
  ggtitle("ITS2 Tree (Short-read)") +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- ggtree(tree_pacbio, aes(color = Phylum)) %<+% pacbio_tax+
  geom_tree(size = 0.2) +
  geom_tiplab(data = function(x) subset(x, label %in% pacbio_labels_to_label),
              aes(label = Genus), size = 4, geom = "text", align = TRUE, linesize = 0.1, show.legend = FALSE) +
  scale_color_manual(values = phylum_colors, na.value = "gray40", name = "Phylum") +
  theme_tree2() +
  ggtitle("Full-length ITS Tree (Long-read)") +
  theme(plot.title = element_text(hjust = 0.5))

# Combine
# Get max distance (branch length) from both trees
max_dist <- 4

# Apply the same x-axis limits to both plots
p1 <- p1 + xlim(0, max_dist)
p2 <- p2 + xlim(0, max_dist)

# Combine side by side
grid.arrange(p1, p2, ncol = 2)
```

###Fig S9: Tree polytomies
```{r}
library(ape)

# Function to count polytomies
count_polytomies <- function(tree) {
  # Count how often each node appears as a parent
  parent_counts <- table(tree$edge[,1])
  
  # Internal nodes only
  internal_nodes <- (Ntip(tree) + 1):(Ntip(tree) + Nnode(tree))
  
  # Count internal nodes with >2 descendants (polytomies)
  sum(parent_counts[as.character(internal_nodes)] > 2, na.rm = TRUE)
}

poly_ITS2  <- count_polytomies(tree_illumina)
poly_full  <- count_polytomies(tree_pacbio)

poly_ITS2
poly_full

#compute proportions
poly_ITS2 / Nnode(tree_illumina)
poly_full / Nnode(tree_pacbio)

#Quantify branch length 
summary(tree_illumina$edge.length)
summary(tree_pacbio$edge.length)

#Distribution statistics
wilcox.test(tree_illumina$edge.length,
            tree_pacbio$edge.length)

#Total tree length
sum(tree_illumina$edge.length)
sum(tree_pacbio$edge.length)

#Visualization: branch length density plot
df_branches <- data.frame(
  length = c(tree_illumina$edge.length,
             tree_pacbio$edge.length),
  Dataset = c(rep("ITS2", length(tree_illumina$edge.length)),
              rep("Full-length ITS", length(tree_pacbio$edge.length)))
)

ggplot(df_branches, aes(x = length, fill = Dataset)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c(
    "ITS2" = "orange",
    "Full-length ITS" = "deeppink"
  )) +
  scale_x_log10() +
  theme_classic() +
  labs(x = "Branch length (log scale)",
       y = "Density")
```


#Compare Blast results: taxonomy
```{r}
# Step 2: Load taxonomy table
tax_ITS <- data.frame(ps_PacBio_ITS_filt@tax_table) 
tax_ITS2 <- data.frame(ps_Illumina_ITS2_filt@tax_table) 
tax_pacbio_ITS2 <- data.frame(ps_PacBio_ITS2_filt@tax_table) 

tax_ITS$Family[grepl("_fam_Incertae_sedis$", tax_ITS$Family)] <- NA

# Ensure the taxonomy table has columns: FeatureID, Taxon
# Step 3: Join top hits with taxonomy

tax_df <- tax_ITS2 %>%
  tibble::rownames_to_column(var = "qseqid")

# Step 5: Merge annotated hits with all ASVs
taxonomy_full <- left_join(tax_df, collapsed_taxonomy, by = "qseqid")
taxonomy_mapped <- subset(taxonomy_full,!is.na(Match_Type))
taxonomy_mapped$Family[grepl("_fam_Incertae_sedis$", taxonomy_mapped$Family)] <- NA

# Step 6: Save output
write.csv(taxonomy_full, "Illumina/illumina_taxonomy_filt.csv")
```

##Fig 5b: Agreement: original taxonomy vs PacBio blasted
```{r}
#Agreement
taxonomy_mapped$Species_Agreement <- ifelse(
  taxonomy_mapped$species == taxonomy_mapped$Species,
  "Match",
  "Mismatch"
)
taxonomy_mapped$Genus_Agreement <- ifelse(
  taxonomy_mapped$genus == taxonomy_mapped$Genus,
  "Match",
  "Mismatch"
)
taxonomy_mapped$Family_Agreement <- ifelse(
  taxonomy_mapped$family == taxonomy_mapped$Family,
  "Match",
  "Mismatch"
)
taxonomy_mapped$Order_Agreement <- ifelse(
  taxonomy_mapped$order == taxonomy_mapped$Order,
  "Match",
  "Mismatch"
)

taxonomy_mapped$Class_Agreement <- ifelse(
  taxonomy_mapped$class == taxonomy_mapped$Class,
  "Match",
  "Mismatch"
)

taxonomy_mapped$Phylum_Agreement <- ifelse(
  taxonomy_mapped$phylum == taxonomy_mapped$Phylum,
  "Match",
  "Mismatch"
)

#Save
write.csv(taxonomy_mapped, "Illumina/illumina_taxonomy_mapped_sameassignmentdatabase_filt.csv")

# Reshape to long format
agreement_long <- taxonomy_mapped %>%
  select(qseqid, Phylum_Agreement, Class_Agreement, Order_Agreement,
         Family_Agreement, Genus_Agreement, Species_Agreement) %>%
  pivot_longer(
    cols = -qseqid,
    names_to = "Rank",
    values_to = "Agreement"
  )

agreement_long <- agreement_long %>%
  mutate(
    Rank = case_when(
      Rank == "Phylum_Agreement"  ~ "Phylum",
      Rank == "Class_Agreement"   ~ "Class",
      Rank == "Order_Agreement"   ~ "Order",
      Rank == "Family_Agreement"  ~ "Family",
      Rank == "Genus_Agreement"   ~ "Genus",
      Rank == "Species_Agreement" ~ "Species",
      TRUE ~ Rank
    )
  )

# Arrange the orders of taxonomy rank
taxonomy_order <- c("Phylum", "Class", "Order", "Family", "Genus","Species")

agreement_long$Rank <- factor(agreement_long$Rank, levels = taxonomy_order)

custom_colors <- c(
  "Match" = "darkgreen",
  "Mismatch" = "red",
  "NA" = "gray"
)

# Plot
Percentage_agreement <- ggplot(agreement_long, aes(x = Rank, fill = Agreement)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = custom_colors) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Taxonomy Assignment Agreement (ITS2 vs Full-length ITS)",
    x = "Taxonomic Rank",
    y = "Percent of ASVs"
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )
Percentage_agreement

breakdown <- agreement_long %>%
  group_by(Rank, Agreement) %>%
  summarise(n = n()) %>%
  group_by(Rank) %>%
  mutate(percent = 100 * n / sum(n))

print(breakdown)
```
##Fig 5c: Improvement in resolution
```{r}
#Step 1: Compute the deepest rank assigned by each method
# Helper function to count non-empty ranks
get_depth <- function(row, ranks) {
  sum(!is.na(row[ranks]) & row[ranks] != "")
}

illumina_ranks <- c("kingdom","phylum", "class", "order", "family", "genus", "species")
pacbio_ranks   <- c("Kingdom","Phylum", "Class", "Order", "Family", "Genus", "Species")

taxonomy_mapped$Illumina_Depth <- apply(taxonomy_mapped, 1, get_depth, ranks = illumina_ranks)
taxonomy_mapped$PacBio_Depth   <- apply(taxonomy_mapped, 1, get_depth, ranks = pacbio_ranks)

#Step 2: Check for full agreement at shared depth
taxonomy_mapped$Agreement_UpTo_Illumina <- mapply(function(row_idx, depth) {
  if (is.na(depth) || depth == 0) return(FALSE)

  illumina_vals <- unlist(taxonomy_mapped[row_idx, illumina_ranks[1:depth]])
  pacbio_vals   <- unlist(taxonomy_mapped[row_idx, pacbio_ranks[1:depth]])

  comparable <- !is.na(illumina_vals) & !is.na(pacbio_vals)

  # If nothing is comparable at all, return NA
  if (!any(comparable)) return(NA)

  all(illumina_vals[comparable] == pacbio_vals[comparable])
},
row_idx = seq_len(nrow(taxonomy_mapped)),
depth = taxonomy_mapped$Illumina_Depth
)

#Step 3: Identify ASVs with improved resolution
taxonomy_mapped$Agreement_Label <- mapply(function(row_idx, depth) {
  if (is.na(depth) || depth == 0) return("No Illumina assignment")

  illumina_vals <- unlist(taxonomy_mapped[row_idx, illumina_ranks[1:depth]])
  pacbio_vals   <- unlist(taxonomy_mapped[row_idx, pacbio_ranks[1:depth]])

  comparable <- !is.na(illumina_vals) & !is.na(pacbio_vals)

  if (!any(comparable)) return("No comparable ranks")
  if (all(illumina_vals[comparable] == pacbio_vals[comparable])) return("Agree")
  return("Disagree")
},
row_idx = seq_len(nrow(taxonomy_mapped)),
depth = taxonomy_mapped$Illumina_Depth
)

taxonomy_mapped$Resolution_Status <- with(taxonomy_mapped, ifelse(
  is.na(PacBio_Depth) | is.na(Illumina_Depth) |
    Agreement_Label %in% c("No Illumina assignment", "No comparable ranks"),
  "Uncomparable",
  ifelse(
    Agreement_Label == "Disagree", "Disagree",
    ifelse(
      PacBio_Depth > Illumina_Depth, "Enhanced",
      ifelse(
        PacBio_Depth == Illumina_Depth, "Same",
        "Decreased"
      )
    )
  )
))

taxonomy_mapped$Resolution_Status <- factor(
  taxonomy_mapped$Resolution_Status,
  levels = c("Same", "Enhanced", "Decreased", "Disagree", "Uncomparable")
)

#Save
write.csv(taxonomy_mapped, "Illumina/illumina_taxonomy_mapped_agreement_filt.csv")

#Step 4: Summarize
Resolution_Change <- ggplot(filter(taxonomy_mapped, !is.na(Resolution_Status) & Resolution_Status != "Uncomparable"), aes(x = Resolution_Status)) +
  geom_bar(fill = "darkgreen", color = "black", linewidth = 0.2) +
  labs(
    title = "Resolution Change of ITS2 ASVs",
    x = "Category",
    y = "Number of ASVs"
  )+
   geom_text(
    stat = "count",
    aes(label = ..count..),
    vjust = 1.5,
    size = 4, color = "white"
  ) +
  theme_bw(base_size = 14) +
   theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )

Resolution_Change
```

##Fig 5e: Missmatched ones
```{r}
# Extract mismatched rows (excluding NAs)
mismatched <- taxonomy_mapped %>%
  filter(Resolution_Status == "Disagree") 

mismatched$assignment.1 <- ifelse(!is.na(mismatched$Species), mismatched$Species,
                            ifelse(!is.na(mismatched$Genus), mismatched$Genus,
                                   ifelse(!is.na(mismatched$Family), mismatched$Family,
                                        ifelse(!is.na(mismatched$Order), mismatched$Order,
                                          ifelse(!is.na(mismatched$Class), mismatched$Class, NA)))))

table(mismatched$phylum)   # Illumina
table(mismatched$Phylum)   # PacBio

#Sankey or alluvial plot
library(ggalluvial)
missmatched <- ggplot(mismatched, aes(axis1 = assignment, axis2 = assignment.1)) +
  geom_alluvium(aes(fill = Phylum), width = 1/12) +
  geom_stratum() + geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("ITS2", "Full-length ITS")) +
  #scale_color_manual(values = phylum_colors, na.value = "gray40", name = "Phylum") +
  scale_fill_brewer(palette = "Dark2", na.value = "gray80") +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(title = "Genus-Level Taxonomic Shifts in Mismatched ASVs")

missmatched

#Create pairwise comparison of each rank
taxonomy_mapped$Mismatch_Rank <- apply(taxonomy_mapped, 1, function(row) {
  for (i in seq_along(illumina_ranks)) {
    illu_val <- row[[illumina_ranks[i]]]
    pacb_val <- row[[pacbio_ranks[i]]]
    
    if (!is.na(illu_val) && illu_val != "" &&
        (!identical(illu_val, pacb_val) || is.na(pacb_val))) {
      return(illumina_ranks[i])
    }
  }
  return("Full match")
})

#Investigate the ranks the mismatch fisrt occured.
# Define your preferred order
rank_order <- c("Full match", "phylum", "class", "order", "family", "genus", "species")

taxonomy_mapped$Mismatch_Rank <- factor(taxonomy_mapped$Mismatch_Rank, levels = rank_order)
setdiff(unique(taxonomy_mapped$Mismatch_Rank), rank_order)

# Filter out "Full match"
taxonomy_mismatches_only <- taxonomy_mapped %>%
  filter(Mismatch_Rank != "Full match")

# Plot only the mismatches
mismatched_rank <- ggplot(taxonomy_mismatches_only, aes(x = Mismatch_Rank)) +
  geom_bar(fill = "coral", color = "black", linewidth = 0.2) +
  theme_bw(base_size = 14) +
   theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  ) +
  labs(
    title = "Rank of First Mismatch",
    x = "Taxonomic Rank",
    y = "Number of ASVs"
  )

mismatched_rank 
```

##Fig 5d: Abundance of missmatched ones
```{r}
# Sum ITS2 ASV counts across all samples
otu <- data.frame(otu_table(ps_Illumina_ITS2_filt))
otu$TotalAbundance <- rowSums(otu)
otu$qseqid <- rownames(otu)
mismatched_abund <- inner_join(mismatched, otu, by = "qseqid")

#Plot and summary
summary(mismatched_abund$TotalAbundance)
hist(mismatched_abund$TotalAbundance/1000, breaks = 100,
     main = "Abundance of Mismatched ASVs", xlab = "Read Count (k)")

# Sum full-length ITS ASV counts across all samples
otu <- data.frame(otu_table(ps_PacBio_ITS_filt))
otu$TotalAbundance <- rowSums(otu)
otu$sseqid <- rownames(otu)

tax <- data.frame(tax_table(ps_PacBio_ITS_filt))
tax$sseqid <- rownames(tax)

mismatched_sseqid <- unlist(strsplit(mismatched$matched_sseqids, ";"))

mismatched_pacbio_abund <- otu[otu$sseqid %in% mismatched_sseqid, ]
mismatched_pacbio_abund <- inner_join(tax, mismatched_pacbio_abund, by = "sseqid")

#Plot and summary
summary(mismatched_pacbio_abund$TotalAbundance)
hist(
  mismatched_pacbio_abund$TotalAbundance / 1000,
  breaks = 100,
  main = "Abundance of Full-length ITS ASVs for Mismatched ITS2 ASVs",
  xlab = "Read Count (k)"
)


```

#Fig 4d: 938 ummapped taxa sequence in ITS2 but not full-length ITS
```{r}
ps_Illumina_ITS2_unmapped
unmapped_taxa_ITS2 <- data.frame(ps_Illumina_ITS2_unmapped@tax_table)

table(unmapped_taxa_ITS2$phylum)
table(unmapped_taxa_ITS2$family)
table(unmapped_taxa_ITS2$genus)

# Genus summary
unmatched_phylum_summary <- unmapped_taxa_ITS2 %>%
  mutate(phylum = ifelse(is.na(phylum) | phylum == "", "Unclassified", phylum)) %>%
  group_by(phylum) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  slice_max(n, n = 20)

unmatched_phylum <- ggplot(unmatched_phylum_summary, aes(x = reorder(phylum, n), y = n)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Phyla of Unmapped ASVs",
    x = "Phylum",
    y = "Number of ASVs"
  )

unmatched_phylum

#Abundance of unmatched:
# Join with OTU count table
otu <- data.frame(otu_table(ps_Illumina_ITS2_unmapped))
otu$TotalAbundance <- rowSums(otu)
otu$qseqid <- rownames(otu)

# Histogram
hist(otu$TotalAbundance, breaks = 50,
     main = "Abundance of Unmmapped ITS2 ASVs",
     xlab = "Read Count")

#What samples they are in?
# Plot abundance or prevalence
plot_bar(ps_Illumina_ITS2_unmapped, fill = "phylum") +
  facet_wrap(~SampleType, scale = "free")+
  labs(title = "Abundance of Illumina-only ASVs")+
theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )
```

## Fig 4e: Health-related taxa abundance
```{r}
# Sample list (expand as needed)
health_relevant_genera <- c("Alternaria", "Aspergillus", "Aureobasidium", "Bipolaris", "Bjerkandera", "Blastomyces", "Botrytis", "Candida", "Chaetomium", "Cladosporium", "Clavispora", "Coccidioides",
  "Cryptococcus", "Debaryomyces", "Drechslera", "Epicoccum", "Eurotium", "Fusarium",
  "Histoplasma", "Malassezia", "Mucor", "Penicillium", "Phoma", "Pichia", "Puccinia",
  "Rhizopus", "Rhodotorula", "Saccharomyces", "Scopulariopsis", "Stachybotrys",
  "Stemphylium", "Trichoderma", "Trichophyton", "Ustilago", "Wallemia", "Wickerhamomyces")

unmapped_taxa_ITS2$HealthRelated <- ifelse(
  unmapped_taxa_ITS2$genus %in% health_relevant_genera, "Yes", "No"
)

table(unmapped_taxa_ITS2$HealthRelated)

otu <- otu %>% 
  tibble::rownames_to_column("TaxaID")

unmapped_taxa_ITS2 <- unmapped_taxa_ITS2 %>% 
  tibble::rownames_to_column("TaxaID")

unmapped_abund <- merge(unmapped_taxa_ITS2, otu, by = "TaxaID")

# Sum their abundance
summary_by_health <- unmapped_abund %>%
  filter(HealthRelated == "Yes") %>%
  group_by(genus) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_healthgenus <- ggplot(summary_by_health, aes(x = reorder(genus, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped health-related ITS2 ASVs",
    x = "Genus",
    y = "Total Read Count"
  )

unmapped_abundance_healthgenus

#Funguil of unmapped
summary_by_function <- unmapped_abund %>%
  group_by(trophicMode) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_function <- ggplot(summary_by_function, aes(x = reorder(trophicMode, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped taxa",
    x = "Trophic Mode",
    y = "Total Read Count"
  )

unmapped_abundance_function

#Growth form
summary_by_growthform <- unmapped_abund %>%
  group_by(growthForm) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_growthmode <- ggplot(summary_by_growthform, aes(x = reorder(growthForm, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped taxa",
    x = "Growth Form",
    y = "Total Read Count"
  )

unmapped_abundance_growthmode
```
#Fig 4d: 4061 Unmapped sequence in PacBio ITS
```{r}
ps_PacBio_Fungi_match_unmapped
unmapped_taxa_ITS <- data.frame(ps_PacBio_Fungi_match_unmapped@tax_table)

table(unmapped_taxa_ITS$Phylum)
table(unmapped_taxa_ITS$Family)
table(unmapped_taxa_ITS$Genus)

# Genus summary
unmatched_phylum_summary <- unmapped_taxa_ITS %>%
  mutate(Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", Phylum)) %>%
  group_by(Phylum) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  slice_max(n, n = 20)

unmatched_phylum <- ggplot(unmatched_phylum_summary, aes(x = reorder(Phylum, n), y = n)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Phyla of Unmapped ASVs",
    x = "Phylum",
    y = "Number of ASVs"
  )

unmatched_phylum

#Abundance of unmatched:
# Join with OTU count table
otu <- data.frame(otu_table(ps_PacBio_Fungi_match_unmapped))
otu$TotalAbundance <- rowSums(otu)
otu$sseqid <- rownames(otu)

summary(otu$TotalAbundance)
thresholds <- c(100, 500, 1000, 3000, 5000)
sapply(thresholds, function(x) {
  mean(otu$TotalAbundance < x) * 100
})

# Histogram
hist(log(otu$TotalAbundance), breaks = 100,
     main = "Abundance of Unmmapped full-length ITS ASVs",
     xlab = "log10 Read Count")

#What samples they are in?
# Plot abundance or prevalence
plot_bar(ps_PacBio_Fungi_match_unmapped, fill = "Phylum") +
  facet_wrap(~SampleType, scale = "free")+
  labs(title = "Abundance of Full-length ITS-only ASVs")+
theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )
```

## Fig 4e: Health-related taxa abundance
```{r}
# Sample list (expand as needed)
health_relevant_genera <- c("Alternaria", "Aspergillus", "Aureobasidium", "Bipolaris", "Bjerkandera", "Blastomyces", "Botrytis", "Candida", "Chaetomium", "Cladosporium", "Clavispora", "Coccidioides",
  "Cryptococcus", "Debaryomyces", "Drechslera", "Epicoccum", "Eurotium", "Fusarium",
  "Histoplasma", "Malassezia", "Mucor", "Penicillium", "Phoma", "Pichia", "Puccinia",
  "Rhizopus", "Rhodotorula", "Saccharomyces", "Scopulariopsis", "Stachybotrys",
  "Stemphylium", "Trichoderma", "Trichophyton", "Ustilago", "Wallemia", "Wickerhamomyces")

unmapped_taxa_ITS$HealthRelated <- ifelse(
  unmapped_taxa_ITS$Genus %in% health_relevant_genera, "Yes", "No"
)

table(unmapped_taxa_ITS$HealthRelated)

otu <- otu %>% 
  tibble::rownames_to_column("TaxaID")

unmapped_taxa_ITS <- unmapped_taxa_ITS %>% 
  tibble::rownames_to_column("TaxaID")

unmapped_abund <- merge(unmapped_taxa_ITS, otu, by = "TaxaID")

# Sum their abundance
summary_by_health <- unmapped_abund %>%
  filter(HealthRelated == "Yes") %>%
  group_by(Genus) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_healthgenus <- ggplot(summary_by_health, aes(x = reorder(Genus, ReadSum/1000), y = ReadSum/1000)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped health-related ITS2 ASVs",
    x = "Genus",
    y = "Total Read Count (k)"
  )

unmapped_abundance_healthgenus

#Funguil of unmapped
summary_by_function <- unmapped_abund %>%
  group_by(trophicMode) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_function <- ggplot(summary_by_function, aes(x = reorder(trophicMode, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped taxa",
    x = "Trophic Mode",
    y = "Total Read Count"
  )

unmapped_abundance_function

#Growth form
summary_by_growthform <- unmapped_abund %>%
  group_by(growthForm) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_growthmode <- ggplot(summary_by_growthform, aes(x = reorder(growthForm, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped taxa",
    x = "Growth Form",
    y = "Total Read Count"
  )

unmapped_abundance_growthmode
```

