---
title: "Rcode_Reviewer1_DepthRarefaction"
author: "Mengyi Dong"
date: "2026-08-06"
output: html_document
---
This document is responding to Reviewer 1's comment on rarefaction and even sequencing depth. We performed rarefaction (1000 depth for both pipelines) and reproduced results for alpha-diversity plots and unmapped ASVs.

#Load packages
```{r}
library(phyloseq)
library(FUNGuildR)
library(Biostrings)
library(qiime2R)
library(tidyr)
library(dplyr)
library(ape)
library(vegan)
library(tibble)
library(ggplot2)
library(stringr)
library(tibble)
library(ggpubr)
library(ggExtra)
library(ggalluvial)
```

#Load data
```{r}
#The PacBio fungi dataset with only the samples matched Illumina ITS2 dataset
ps_PacBio_Fungi_match <- readRDS("PacBio/ps_PacBio_Fungi_match_sklearn.rds")
tax_ITS <- data.frame(ps_PacBio_Fungi_match@tax_table)
metadata_ITS <- data.frame(ps_PacBio_Fungi_match@sam_data)
otu_ITS <- data.frame(ps_PacBio_Fungi_match@otu_table)

#The PacBio fungi dataset trimmed to ITS2 region with only samples match Illumina ITS2 dataset
ps_PacBio_ITS2_match <- readRDS("PacBio/ps_PacBio_ITS2_match.rds")
tax_pacbio_ITS2 <- data.frame(ps_PacBio_ITS2_match@tax_table)
metadata_pacbio_ITS2 <- data.frame(ps_PacBio_ITS2_match@sam_data)
otu_pacbio_ITS2 <- data.frame(ps_PacBio_ITS2_match@otu_table)

#The cleaned Illumina ITS2 dataset
ps_Illumina_ITS2 <- readRDS("Illumina/ps_Illumina_ITS2_sklearn.rds")
tax_ITS2 <- data.frame(ps_Illumina_ITS2@tax_table)
metadata_ITS2 <- data.frame(ps_Illumina_ITS2@sam_data)
otu_ITS2 <- data.frame(ps_Illumina_ITS2@otu_table)
```

#Rarefaction
##Fig S2: determine where a plateau occurs for each dataset
```{r}
# Convert to matrix, make sure OTUs are in rows for vegan
otu_ITS <- t(as(otu_table(ps_PacBio_Fungi_match), "matrix"))
otu_ITS2 <- t(as(otu_table(ps_Illumina_ITS2), "matrix"))
otu_ITS2_pacbio <- t(as(otu_table(ps_PacBio_ITS2_match), "matrix"))

otu_ITS2_filt <- otu_ITS2[rowSums(otu_ITS2, na.rm = TRUE) > 1, , drop = FALSE]

# Plot rarefaction curves
rarecurve(otu_ITS, step = 1000, sample=min(rowSums(otu_ITS)), label = FALSE, col = "deeppink", main = "PacBio ITS Rarefaction")
rarecurve(otu_ITS2_filt, step = 1000, sample=min(rowSums(otu_ITS2)),label = FALSE, col = "orange", main = "Illumina ITS2 Rarefaction")
rarecurve(otu_ITS2_pacbio, step = 1000, sample=min(rowSums(otu_ITS2_pacbio)),label = FALSE, col = "deeppink", main = "Trimmed ITS2 Rarefaction")
```

##Choose a Rarefaction Depth
```{r}
summary(sample_sums(ps_PacBio_Fungi_match))
summary(sample_sums(ps_Illumina_ITS2))

depth_shared <- 1000  # Choose based on previous rarefaction curve

# Rarefy both to same depth
set.seed(123)
ps_PacBio_rarefied <- rarefy_even_depth(ps_PacBio_Fungi_match, sample.size = depth_shared, rngseed = 123, verbose = FALSE)
ps_Illumina_rarefied <- rarefy_even_depth(ps_Illumina_ITS2, sample.size = depth_shared, rngseed = 123, verbose = FALSE)

ps_PacBio_rarefied
ps_Illumina_rarefied

# Number of non-zero taxa per sample = ASVs detected
asv_counts_PacBio <- apply(otu_table(ps_PacBio_rarefied), 2, function(x) sum(x > 0))
asv_counts_Illumina <- apply(otu_table(ps_Illumina_rarefied), 2, function(x) sum(x > 0))

# Convert both vectors to data frames
common_samples <- intersect(names(asv_counts_PacBio), names(asv_counts_Illumina))


asv_df <- data.frame(
  SampleID = common_samples,
  ASVs_PacBio = asv_counts_PacBio[common_samples],
  ASVs_Illumina = asv_counts_Illumina[common_samples]
)

#Plot for comparison
summary_ASV_detected <- ggplot(asv_df, aes(x = ASVs_Illumina, y = ASVs_PacBio)) +
  geom_point(size = 3, alpha = 0.3, color = "black") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )+
  labs(x = "ASVs (Short-read)", y = "ASVs (Long-read)", title = paste("ASV Richness Comparison at", depth_shared, "Reads"))

summary_ASV_detected

#Visualize paired difference
asv_long <- asv_df %>%
  pivot_longer(cols = starts_with("ASVs"), names_to = "Pipeline", values_to = "ASV_Count")

paired <- ggplot(asv_long, aes(x = Pipeline, y = ASV_Count, group = SampleID)) +
  geom_line(alpha = 0.3) +
  geom_point(size = 3, alpha = 0.3) +
  labs(title = "ASVs Detected per Sample by Pipeline (rarefied to 5000)", y = "Number of ASVs", x = "Pipeline") +
  theme_bw(base_size = 14)+
   theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )

paired
```

#Subset common samples
```{r}
common_samples <- intersect(
  sample_names(ps_PacBio_rarefied),
  sample_names(ps_Illumina_rarefied)
)

length(common_samples)

ps_PacBio_pair <- prune_samples(
  common_samples,
  ps_PacBio_rarefied
)

ps_Illumina_pair <- prune_samples(
  common_samples,
  ps_Illumina_rarefied
)

ps_PacBio_pair <- prune_taxa(
  taxa_sums(ps_PacBio_pair) > 0,
  ps_PacBio_pair
)

ps_Illumina_pair <- prune_taxa(
  taxa_sums(ps_Illumina_pair) > 0,
  ps_Illumina_pair
)
```


#Alpha-diversity group by sample type
```{r}
# 1. Compute alpha diversity for each pipeline
rarefied_ITS2_alpha <- estimate_richness(ps_Illumina_pair,
                                measures = c("Shannon", "Chao1", "Observed")) %>%
  mutate(Label = rownames(.),
         Pipeline = "ITS2")


rarefied_ITS_alpha <- estimate_richness(ps_PacBio_pair,
                                   measures = c("Shannon", "Chao1", "Observed")) %>%
  mutate(Label = rownames(.),
         Pipeline = "Full-length ITS")

# 2. Inner join on Label to ensure true pairs
alpha_pairs <- inner_join(
  rarefied_ITS2_alpha %>% select(Label, Shannon_ITS2 = Shannon,
                        Chao1_ITS2 = Chao1, Obs_ITS2 = Observed),
  rarefied_ITS_alpha %>% select(Label, Shannon_ITS = Shannon,
                           Chao1_ITS = Chao1, Obs_ITS = Observed),
  by = "Label"
)

metadata <- data.frame(ps_Illumina_pair@sam_data)
alpha_pairs <- left_join(alpha_pairs, metadata, by = "Label")

# Reshape the dataframe
alpha_long <- alpha_pairs %>%
  transmute(
    Label, SampleType, Location,
    Shannon_ITS2 = Shannon_ITS2, Shannon_ITS = Shannon_ITS,
    Chao1_ITS2   = Chao1_ITS2,   Chao1_ITS   = Chao1_ITS,
    Observed_ITS2 = Obs_ITS2,    Observed_ITS = Obs_ITS
  ) %>%
  pivot_longer(
    cols = -c(Label, SampleType, Location),
    names_to  = c("Metric","Pipeline"),
    names_sep = "_",
    values_to = "Value"
  ) %>%
  mutate(Pipeline = ifelse(Pipeline == "ITS2", "ITS2", "Full-length ITS"))

# 3. Paired Wilcoxon tests
paired_alpha_tests <- alpha_long %>%
  pivot_wider(id_cols = c(Label, SampleType, Metric),
              names_from = Pipeline,
              values_from = Value) %>%
  filter(!is.na(`Full-length ITS`), !is.na(ITS2)) %>%
  group_by(SampleType, Metric) %>%
  summarise(
    n           = n(),
    median_diff = median(`Full-length ITS` - ITS2, na.rm = TRUE),
    p_value     = wilcox.test(`Full-length ITS`, ITS2, paired = TRUE,
                              exact = FALSE)$p.value,
    W           = wilcox.test(`Full-length ITS`, ITS2, paired = TRUE,
                              exact = FALSE)$statistic,
    .groups = "drop"
  ) %>%
  group_by(Metric) %>%                # FDR across SampleTypes within each metric
  mutate(q_value = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  mutate(sig = case_when(
    is.na(q_value)      ~ "",
    q_value < 0.001     ~ "***",
    q_value < 0.01      ~ "**",
    q_value < 0.05      ~ "*",
    TRUE                ~ "ns"
  ))

#4. Plot: paired box with lines
desired_location_order <- c("Control","Dorm", "Home1", "Home2", "Lab1", "Lab2")
alpha_long$Location <- factor(alpha_long$Location, levels = desired_location_order)

plot_alpha_paired_facet_by_loc <- function(metric_name, ylab) {

  df_plot <- alpha_long %>% 
    filter(Metric == metric_name) %>%
    # ensure Location exists (pulled from your metadata join)
    mutate(Location = factor(Location))      

  # y-position per facet (just above max in that facet)
  y_max <- df_plot %>%
    group_by(SampleType) %>%
    summarise(y = max(Value, na.rm = TRUE), .groups = "drop")

  ann <- paired_alpha_tests %>%
    filter(Metric == metric_name) %>%
    select(SampleType, sig, q_value) %>%
    left_join(y_max, by = "SampleType") %>%
    mutate(y = y + 0.05 * abs(y),   # headroom for the annotation
           x = 1.5)

  ggplot(df_plot, aes(x = Pipeline, y = Value, group = Label)) +
     # side-by-side boxplots per Pipeline (keep neutral fill)
    geom_boxplot(aes(group = Pipeline),
                 width = 0.6, fill = "grey88", color = "black",
                 outlier.shape = NA) +
    # neutral connecting lines emphasize pairing (not method or location)
    geom_line(alpha = 0.35, color = "grey55") +
    # points colored by Location (Dark2)
    geom_point(aes(color = Location), size = 2, alpha = 0.9,
               position = position_jitter(width = 0.04, height = 0)) +
    # significance label per facet (paired Wilcoxon within SampleType)
    geom_text(data = ann, aes(x = x, y = y, label = sig),
              inherit.aes = FALSE, size = 5) +
    scale_color_brewer(palette = "Set2", drop = FALSE) +  # <-- Dark2
    facet_wrap(~ SampleType, scales = "fixed") +
    labs(x = "", y = ylab,
         color = "Location",
         caption = "Paired Wilcoxon within SampleType; labels: ***<0.001, **<0.01, *<0.05, ns≥0.05") +
    theme_classic(base_size = 14) +
    theme(
      panel.border = element_rect(color = "black", fill = NA),
      axis.text.x  = element_text(angle = 45, hjust = 1)
    )
}

# Build your three panels
p_shannon  <- plot_alpha_paired_facet_by_loc("Shannon",  "Shannon diversity")
p_chao1    <- plot_alpha_paired_facet_by_loc("Chao1",    "Chao1 richness")
p_observed <- plot_alpha_paired_facet_by_loc("Observed", "Observed ASVs")

p_shannon; p_chao1; p_observed
```


#Unmapped ASVs
```{r}
# Load libraries
library(dplyr)
library(readr)

# Step 1: Load BLAST hits
blast_hits <- read_tsv("Illumina/Illumina_blast_hits_allpacbio.tsv", col_names = c(
  "qseqid", "sseqid", "pident", "length", "mismatch", "gapopen",
  "qstart", "qend", "sstart", "send", "evalue", "bitscore"
))

seq_asv_map <- tibble(
  SequenceID = names(refseq(ps_Illumina_pair)),
  Sequence = as.character(refseq(ps_Illumina_pair))
)

# Extract taxonomy table
refseqs <- refseq(ps_PacBio_pair)
tax <- tax_table(ps_PacBio_pair)

# Convert to data frame and add sequence IDs
tax_df_ITS <- as.data.frame(tax)
tax_df_ITS$SequenceID <- names(refseqs)

blast_hits_filt <- blast_hits %>%
  filter(
    qseqid %in% seq_asv_map$Sequence,
    sseqid %in% tax_df_ITS$SequenceID
  )

blast_hits_filt$qseqid_sequence <-blast_hits_filt$qseqid
blast_hits_filt$qseqid <- seq_asv_map$SequenceID[
  match(blast_hits_filt$qseqid_sequence, seq_asv_map$Sequence)
]

#Mapped hits
matched_hits <- blast_hits_filt %>%
  filter(pident >= 99, bitscore > 0)

matched_summary <- matched_hits %>%
  count(qseqid, sseqid)

#Extract all 100% hits
hits_100 <- matched_hits %>%
  filter(pident >= 99.999)
```

##Subset to mapped and unmapped phyloseq
```{r}
ps_Illumina_ITS2_mapped <- subset_taxa(ps_Illumina_pair, taxa_names(ps_Illumina_pair) %in% hits_100$qseqid)
ps_Illumina_ITS2_unmapped <- subset_taxa(ps_Illumina_pair, !(taxa_names(ps_Illumina_pair) %in% hits_100$qseqid))

ps_PacBio_Fungi_match_mapped <- subset_taxa(ps_PacBio_pair, taxa_names(ps_PacBio_pair) %in% hits_100$sseqid)
ps_PacBio_Fungi_match_unmapped <- subset_taxa(ps_PacBio_pair, !(taxa_names(ps_PacBio_pair) %in% hits_100$sseqid))

ps_Illumina_ITS2_mapped
ps_Illumina_ITS2_unmapped
ps_PacBio_Fungi_match_mapped
ps_PacBio_Fungi_match_unmapped

Illumina_ITS2_mapped_tax <- data.frame(ps_Illumina_ITS2_mapped@tax_table)
PacBio_ITS_mapped_tax <- data.frame(ps_PacBio_Fungi_match_mapped@tax_table)
```

# 436 ummapped taxa sequence in ITS2 but not full-length ITS
```{r}
ps_Illumina_ITS2_unmapped
unmapped_taxa_ITS2 <- data.frame(ps_Illumina_ITS2_unmapped@tax_table)

table(unmapped_taxa_ITS2$phylum)
table(unmapped_taxa_ITS2$family)
table(unmapped_taxa_ITS2$genus)

# Genus summary
unmapped_phylum_summary <- unmapped_taxa_ITS2 %>%
  mutate(phylum = ifelse(is.na(phylum) | phylum == "", "Unclassified", phylum)) %>%
  group_by(phylum) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  slice_max(n, n = 20)

unmapped_phylum <- ggplot(unmapped_phylum_summary, aes(x = reorder(phylum, n), y = n)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Phyla of Unmapped ASVs",
    x = "Phylum",
    y = "Number of ASVs"
  )

unmapped_phylum

#Abundance of unmatched:
# Join with OTU count table
otu <- data.frame(otu_table(ps_Illumina_ITS2_unmapped))
otu$TotalAbundance <- rowSums(otu)
otu$qseqid <- rownames(otu)

# Histogram
hist(otu$TotalAbundance, breaks = 50,
     main = "Abundance of Unmmapped ITS2 ASVs",
     xlab = "Read Count")

#What samples they are in?
# Plot abundance or prevalence
plot_bar(ps_Illumina_ITS2_unmapped, fill = "phylum") +
  facet_wrap(~SampleType, scale = "free")+
  labs(title = "Abundance of Illumina-only ASVs")+
theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )
```

## Health-related taxa abundance: ITS2
```{r}
# Sample list (expand as needed)
health_relevant_genera <- c("Alternaria", "Aspergillus", "Aureobasidium", "Bipolaris", "Bjerkandera", "Blastomyces", "Botrytis", "Candida", "Chaetomium", "Cladosporium", "Clavispora", "Coccidioides",
  "Cryptococcus", "Debaryomyces", "Drechslera", "Epicoccum", "Eurotium", "Fusarium",
  "Histoplasma", "Malassezia", "Mucor", "Penicillium", "Phoma", "Pichia", "Puccinia",
  "Rhizopus", "Rhodotorula", "Saccharomyces", "Scopulariopsis", "Stachybotrys",
  "Stemphylium", "Trichoderma", "Trichophyton", "Ustilago", "Wallemia", "Wickerhamomyces")

unmapped_taxa_ITS2$HealthRelated <- ifelse(
  unmapped_taxa_ITS2$genus %in% health_relevant_genera, "Yes", "No"
)

table(unmapped_taxa_ITS2$HealthRelated)

otu <- otu %>% 
  tibble::rownames_to_column("TaxaID")

unmapped_taxa_ITS2 <- unmapped_taxa_ITS2 %>% 
  tibble::rownames_to_column("TaxaID")

unmapped_abund <- merge(unmapped_taxa_ITS2, otu, by = "TaxaID")

# Sum their abundance
summary_by_health <- unmapped_abund %>%
  filter(HealthRelated == "Yes") %>%
  group_by(genus) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_healthgenus <- ggplot(summary_by_health, aes(x = reorder(genus, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped health-related ITS2 ASVs",
    x = "Genus",
    y = "Total Read Count"
  )

unmapped_abundance_healthgenus
```

##Funguil of unmapped: ITS2
```{r}
#By function
summary_by_function <- unmapped_abund %>%
  group_by(trophicMode) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_function <- ggplot(summary_by_function, aes(x = reorder(trophicMode, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped taxa",
    x = "Trophic Mode",
    y = "Total Read Count"
  )

unmapped_abundance_function

#Growth form
summary_by_growthform <- unmapped_abund %>%
  group_by(growthForm) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_growthmode <- ggplot(summary_by_growthform, aes(x = reorder(growthForm, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "orange", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped taxa",
    x = "Growth Form",
    y = "Total Read Count"
  )

unmapped_abundance_growthmode
```

# 1392 Unmapped sequence in PacBio ITS
```{r}
ps_PacBio_Fungi_match_unmapped
unmapped_taxa_ITS <- data.frame(ps_PacBio_Fungi_match_unmapped@tax_table)

table(unmapped_taxa_ITS$Phylum)
table(unmapped_taxa_ITS$Family)
table(unmapped_taxa_ITS$Genus)

# Genus summary
unmatched_phylum_summary <- unmapped_taxa_ITS %>%
  mutate(Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", Phylum)) %>%
  group_by(Phylum) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  slice_max(n, n = 20)

unmatched_phylum <- ggplot(unmatched_phylum_summary, aes(x = reorder(Phylum, n), y = n)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Phyla of Unmapped ASVs",
    x = "Phylum",
    y = "Number of ASVs"
  )

unmatched_phylum

#Abundance of unmatched:
# Join with OTU count table
otu <- data.frame(otu_table(ps_PacBio_Fungi_match_unmapped))
otu$TotalAbundance <- rowSums(otu)
otu$sseqid <- rownames(otu)

summary(otu$TotalAbundance)
thresholds <- c(100, 500, 1000, 3000, 5000)
sapply(thresholds, function(x) {
  mean(otu$TotalAbundance < x) * 100
})

# Histogram
hist(log(otu$TotalAbundance), breaks = 100,
     main = "Abundance of Unmmapped full-length ITS ASVs",
     xlab = "log10 Read Count")

#What samples they are in?
# Plot abundance or prevalence
plot_bar(ps_PacBio_Fungi_match_unmapped, fill = "Phylum") +
  facet_wrap(~SampleType, scale = "free")+
  labs(title = "Abundance of Full-length ITS-only ASVs")+
theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1)  # Rotate x-axis labels
  )
```

## Health-related taxa abundance: full-length ITS
```{r}
# Sample list (expand as needed)
health_relevant_genera <- c("Alternaria", "Aspergillus", "Aureobasidium", "Bipolaris", "Bjerkandera", "Blastomyces", "Botrytis", "Candida", "Chaetomium", "Cladosporium", "Clavispora", "Coccidioides",
  "Cryptococcus", "Debaryomyces", "Drechslera", "Epicoccum", "Eurotium", "Fusarium",
  "Histoplasma", "Malassezia", "Mucor", "Penicillium", "Phoma", "Pichia", "Puccinia",
  "Rhizopus", "Rhodotorula", "Saccharomyces", "Scopulariopsis", "Stachybotrys",
  "Stemphylium", "Trichoderma", "Trichophyton", "Ustilago", "Wallemia", "Wickerhamomyces")

unmapped_taxa_ITS$HealthRelated <- ifelse(
  unmapped_taxa_ITS$Genus %in% health_relevant_genera, "Yes", "No"
)

table(unmapped_taxa_ITS$HealthRelated)

otu <- otu %>% 
  tibble::rownames_to_column("TaxaID")

unmapped_taxa_ITS <- unmapped_taxa_ITS %>% 
  tibble::rownames_to_column("TaxaID")

unmapped_abund <- merge(unmapped_taxa_ITS, otu, by = "TaxaID")

# Sum their abundance
summary_by_health <- unmapped_abund %>%
  filter(HealthRelated == "Yes") %>%
  group_by(Genus) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_healthgenus <- ggplot(summary_by_health, aes(x = reorder(Genus, ReadSum/1000), y = ReadSum/1000)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped health-related ITS2 ASVs",
    x = "Genus",
    y = "Total Read Count (k)"
  )

unmapped_abundance_healthgenus
```

##Funguil of unmapped: full-length ITS
```{r}
#By function
summary_by_function <- unmapped_abund %>%
  group_by(trophicMode) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_function <- ggplot(summary_by_function, aes(x = reorder(trophicMode, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped taxa",
    x = "Trophic Mode",
    y = "Total Read Count"
  )

unmapped_abundance_function

#Growth form
summary_by_growthform <- unmapped_abund %>%
  group_by(growthForm) %>%
  summarize(ReadSum = sum(TotalAbundance)) %>%
  arrange(desc(ReadSum))

unmapped_abundance_growthmode <- ggplot(summary_by_growthform, aes(x = reorder(growthForm, ReadSum), y = ReadSum)) +
  geom_bar(stat = "identity", fill = "deeppink", color = "black", linewidth = 0.2) +
  coord_flip() +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),  # Remove major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
  )+
  labs(
    title = "Abundance of unmapped taxa",
    x = "Growth Form",
    y = "Total Read Count"
  )

unmapped_abundance_growthmode
```
